---
title: "Untitled"
output: html_document
date: "2026-01-21"
---

```{r setup, set.seed(711), include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preliminaries

## Libraries

```{r}
library(devtools)

## 1. Making phyloseq
library(dplyr)
library(tidyverse)
library("phyloseq")
library("microbiome")
#devtools::install_github("benjjneb/decontam")
library("decontam")
library(ggplot2)
library(stringr)


## 2
library("phyloseq.extended")

## 3
library(doBy)
library(vegan)
library(microbiome)
library(RVAideMemoire)
library(FSA)

#3.1
library(RColorBrewer)
library(scales)
library(colorspace)
library(ANCOMBC)
library(microbiomeMarker)
library(DESeq2)

#15
library(viridis)
library(scales)
```

## Theme for graphs

```{r}
mytheme <- theme_bw()+
  theme(text = element_text(size=15),
        strip.background = element_blank(),
        strip.placement = "outside",
        legend.box.background = element_rect(),
        legend.box.margin = margin(1, 1, 1, 1),
        legend.title = element_text(face = "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())
```

## Functions

```{r}
# Input: 
#data = df, where variable1,variable2 and value need to be columnames. 
#variable1: What do you wanna look at, e.g. typology. 
#variable2: The thing you want to compare. so if var1 is typology, you can compare them by var2 (e.g. city)
#Value: Richness/shannon/etc.


compare_variables_function <- function(data, variable1, variable2, value) {
  statement_return <- list()
  # Loop through each level of the first grouping variable
  for (subgroup in unique(data[[variable1]])) {
    
    # Subset data for the current subgroup based on variable1
    subgroup_data <- data[data[[variable1]] == subgroup, ]
    
    # Check if subgroup_data is empty
    if (nrow(subgroup_data) == 0) {
      next
    }
    # Perform normality test (Shapiro-Wilk)
    normality_test <- shapiro.test(as.numeric(subgroup_data[[value]]))
    
    # Perform Bartlett test for homogeneity of variances
    bartlett_test <- bartlett.test(subgroup_data[[value]], subgroup_data[[variable2]])
    
    if (normality_test$p.value > 0.05 & bartlett_test$p.value > 0.05) {
      # If both normality and homogeneity are met, perform ANOVA
      anova_result <- aov(subgroup_data[[value]] ~ subgroup_data[[variable2]])
      tukey_result <- TukeyHSD(anova_result)
      statement_return[[subgroup]] <- list(anova_result = anova_result, tukey_result = tukey_result)
    } else {
      # If either normality or homogeneity is violated, perform non-parametric test
      kruskal_result <- kruskal.test(subgroup_data[[value]] ~ subgroup_data[[variable2]])
      wilcox_result <- pairwise.wilcox.test(subgroup_data[[value]], subgroup_data[[variable2]], p.adjust.method = "BH")
      #Added dunntest because that is the one that is preferably used
      dunntest_result <- dunnTest(subgroup_data[[value]] ~ subgroup_data[[variable2]], method = "bh")
      statement_return[[subgroup]] <- list(kruskal_result = kruskal_result, wilcox_result = wilcox_result, dunntest_result = dunntest_result)
    }
  }
  
  return(statement_return)
}
```

### Taxonomy Bar Graph function

```{r}
# Input should be presence absence physeq
plot_grouped_taxa_bar <- function(ps_obj, 
                                  merge_var,
                                  level = "Class", 
                                  higher_level = "Phylum", 
                                  threshold = 3.5,
                                  palette = "Set2") {
  
   pretty_names <- function(x) {
    s <- gsub("_", " ", x)
    tools::toTitleCase(s)
  }
  
  # Create a list to control order of bars
   order_graphs <- unique(sample_data(ps_obj)[[merge_var]])
   
   
  # Merge and normalize
  ps_merged <- merge_samples(ps_obj, merge_var)
  ps_rel <- transform_sample_counts(ps_merged, function(x) x / sum(x) * 100)
  ps_tax <- tax_glom(ps_rel, taxrank = level, NArm = FALSE)
  
  # Clean taxonomy
  taxdf <- as.data.frame(tax_table(ps_tax))
  taxdf[[level]] <- as.character(taxdf[[level]])
  taxdf[[higher_level]] <- as.character(taxdf[[higher_level]])
  
  ambiguous <- is.na(taxdf[[higher_level]]) |
    tolower(taxdf[[higher_level]]) == "unknown" |
    grepl("incertae", taxdf[[higher_level]], ignore.case = TRUE) |
    grepl("unclassified", taxdf[[higher_level]], ignore.case = TRUE)
  taxdf[[higher_level]][ambiguous] <- "Unclassified"
  taxdf[[level]][ambiguous] <- "Unclassified"
  
  # Collapse low-abundance phyla into a single group
  abund_phylum <- tapply(taxa_sums(ps_tax), taxdf[[higher_level]], sum)
  rel_abund_phylum <- abund_phylum / sum(abund_phylum) * 100
  low_phyla <- names(rel_abund_phylum[rel_abund_phylum < threshold])
  taxdf[[higher_level]][taxdf[[higher_level]] %in% low_phyla] <- "Other_Phyla"
  taxdf[[level]][taxdf[[higher_level]] == "Other_Phyla"] <- "Other_Phyla"
  
  # n Collapse low-abundance classes (only for remaining phyla)
  abund <- taxa_sums(ps_tax) / sum(taxa_sums(ps_tax)) * 100
  low_taxa <- setdiff(names(abund[abund < threshold]),
                      rownames(taxdf)[taxdf[[level]] %in% c("Unclassified", "Other_Phyla")])
  
  taxdf[[level]][rownames(taxdf) %in% low_taxa] <- paste0("Other_", taxdf[[higher_level]][rownames(taxdf) %in% low_taxa])
  # Collapse and reapply taxonomy
  tax_table(ps_tax) <- tax_table(as.matrix(taxdf))
  ps_tax <- tax_glom(ps_tax, taxrank = level)
  taxdf <- as.data.frame(tax_table(ps_tax))
  
  # Order phyla and classes by abundance
  abund_df <- psmelt(ps_tax)
  abund_summary <- aggregate(Abundance ~ get(higher_level) + get(level), data = abund_df, sum)
  colnames(abund_summary) <- c(higher_level, level, "Abundance")
  phyla_order <- aggregate(Abundance ~ get(higher_level), data = abund_summary, sum)
  colnames(phyla_order) <- c(higher_level, "Total")
  phyla_order <- phyla_order[order(-phyla_order$Total), ]
  
  taxdf[[higher_level]] <- factor(taxdf[[higher_level]], levels = phyla_order[[higher_level]])
  taxdf[[level]] <- factor(taxdf[[level]], levels = abund_summary[order(
    match(abund_summary[[higher_level]], phyla_order[[higher_level]]),
    -abund_summary$Abundance
  ), level])
  tax_table(ps_tax) <- tax_table(as.matrix(taxdf))
  
  # Color palette - use shuffled order just for color assignment
  phyla_ordered <- levels(taxdf[[higher_level]])
  phyla_shuffled <- phyla_ordered[order(nchar(phyla_ordered), phyla_ordered)]
  
  base_colors <- brewer.pal(min(length(phyla_shuffled), 8), palette)
  if (length(phyla_shuffled) > length(base_colors)) {
    base_colors <- colorRampPalette(brewer.pal(8, palette))(length(phyla_shuffled))
  }
  phylum_colors <- setNames(base_colors, phyla_shuffled)
  
  # Shades for class-level coloring (using original phylum order for logic)
  class_colors <- c()
  for (ph in phyla_ordered) {
    classes <- unique(taxdf[[level]][taxdf[[higher_level]] == ph])
    n <- length(classes)
    if (n > 0) {
      base_col <- phylum_colors[ph]
      shades <- colorRampPalette(c("gray90", base_col, "black"))(n + 4)[2:(n + 1)]
      class_colors <- c(class_colors, setNames(shades, classes))
    }
  }
  
  # Legend labels
  legend_labels <- setNames(
    pretty_names(paste0(names(class_colors), " (", taxdf[[higher_level]][match(names(class_colors), taxdf[[level]])], ")")),
    names(class_colors)
  )
  
  # Plot
  p <- plot_bar(ps_tax, fill = level, title = paste("Barplot of ",level," with abundance of more than ", threshold,"%"))
  p$data[[level]] <- factor(p$data[[level]], levels = levels(taxdf[[level]]))
  
  # Remove the original geom_bar and add a new one with no border
  p$layers <- NULL  # remove default geom_bar layer
  
  p <- p +
    scale_x_discrete(limits = order_graphs) +
    geom_bar(stat = "identity", position = "stack", color = NA) +  # <- no borders
    scale_fill_manual(values = class_colors, labels = legend_labels) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.title = element_blank(),
          legend.text = element_text(size = 9)) +
    labs(y = "Relative abundance (%)")
  return(p)
}

```

# ITS

## 0.1 Decontam ITS phyloseq

```{r}
otu_table=read.csv("G:/My Drive/CityDeal/Data_analysis_after_review/Input_data/ITS_spring23_ASV_Unrarefied_simplefied.csv",row.names=1, header = T)
head(otu_table)
otu_table=as.matrix(otu_table)
mean(colSums(otu_table))
nrow(otu_table)

taxonomy=read.csv("G:/My Drive/CityDeal/Data_analysis_after_review/Input_data/080425-Taxa_list_all_eurkaryotes_and_ASVs.csv", row.names=1, header = T)
taxonomy$sequence <- NULL
replace(taxonomy, taxonomy == " ", NA) -> tax1
taxonomy1=as.matrix(tax1)

## Metadata
metadata=read.csv("./Input_data/FA_metadata_ITS.csv", row.names=1)

rownames(metadata) <- metadata$SampleID

TAX = tax_table(as.matrix(taxonomy1))
OTU=otu_table(otu_table, taxa_are_rows=TRUE)
META=sample_data(metadata)
physeq_ITS_full_set=phyloseq(OTU, META, TAX)

```

```{r}
#settings
theme_set(theme_bw())
options(stringsAsFactors = FALSE)
path.in <- "G:/My Drive/CityDeal/Data_analysis_after_review/Input_data" # CHANGE ME
path.out <- "G:/My Drive/CityDeal/Data_analysis_after_review/Decntam" # CHANGE ME

#color palettes
cbbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7","#000000")
methodPalette <- c(Frequency="darkorange2", Prevalence="turquoise3", Combined="darkorchid4")
prevalencePalette <- c("2" = "#edf8e9", "3-5" = "#bae4b3", "6-10" = "#74c476", "11+" = "#238b45")

#load phyloseq object
load(file.path(path.in, "physeq_ITS_full_set"))
physeq_ITS_full_set
```

Classify contaminants

```{r}
conc2 <- physeq_ITS_full_set@sam_data$ng.uL # ng/ul
neg2 <- physeq_ITS_full_set@sam_data$sample_control == 'neg_control' # True if a negative control sample
# Frequency-based contaminant classification
ocf <- isContaminant(physeq_ITS_full_set, conc=conc2, threshold=0.1, detailed=TRUE, normalize=TRUE, method='frequency')
# Prevalence-based contaminant classification
ocp <- isContaminant(physeq_ITS_full_set, neg=neg2, threshold=0.1, detailed=TRUE, normalize=TRUE, method='prevalence')
# Combined contaminant classification
occ <- isContaminant(physeq_ITS_full_set, conc=conc2, neg=neg2, threshold=0.1, detailed=TRUE, normalize=TRUE, method='combined')
```

```{r}
probcols <- data.frame(row.names=rownames(ocf),prob.f=ocf$p.freq, prob.p=ocp$p.prev, prob.c = occ$p)

TAX2 <- as.matrix(physeq_ITS_full_set@tax_table)
TAX2 <- cbind(TAX2, probcols[rownames(TAX2),])
### MANY ASVs have NA, these all occur in only one sample. I will not remove them, because frankly we can be quite sure they are no contaminants
# Crude comparison of frequency and prevalence contaminant assignment
table(probcols$prob.f<0.1, probcols$prob.c<0.1)
```

bar plots

```{r}
# Augment with prevalence groups and Abundance column
physeq_ITS_prev <- transform_sample_counts(physeq_ITS_full_set, function(x) ifelse(x>0, 1, 0))
TAX2$TAXprev <- taxa_sums(physeq_ITS_prev)[rownames(TAX2)] # Add prevalences
TAX2$Prevalence <- cut(TAX2$TAXprev, c(0, 2, 5, 10, 9999), labels=c("2", "3-5", "6-10", "11+"))

TAXann <- rbind(cbind(TAX2, Score=TAX2$prob.f, Method="Frequency"),
                cbind(TAX2, Score=TAX2$prob.p, Method="Prevalence"),
                cbind(TAX2, Score=TAX2$prob.c, Method="Combined"))

TAXann$Method <- factor(TAXann$Method, levels=c('Frequency','Prevalence','Combined'))
prevalencePalette <- c("2" = "#edf8e9", "3-5" = "#bae4b3", "6-10" = "#74c476", "11+" = "#238b45")



TAXannTest <- subset(TAXann, Method != "Prevalence")
#figure 2b
histo <- ggplot(TAXannTest, aes(x=Score, fill=Prevalence))
histo
histo <- histo + geom_histogram() + labs(x = 'decontam Score', y='Number ASVs') + 
  facet_wrap(~Method, nrow=1) +
  #http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=4
  scale_fill_manual(values=prevalencePalette) +
  scale_y_sqrt()+
 # theme(axis.text.x = element_text(colour="grey20",size=10,angle=0,
 #                                   hjust=.5,vjust=.5,face="plain"),
 #       axis.text.y = element_text(colour="grey20",size=10,angle=0,
 #                                  hjust=1,vjust=0,face="plain"),
 #       axis.title.x = element_text(colour="grey20",size=15,angle=0,
 #                                   hjust=.5,vjust=0,face="plain"),
 #       axis.title.y = element_text(colour="grey20",size=15,angle=90,
 #                                   hjust=.5,vjust=.5,face="plain"),
 #       strip.text.x = element_text(size = 15, angle = 0),
 #       legend.title = element_text(size = 15, angle = 0),
 #       legend.text = element_text(size = 10, angle = 0)) +
  theme(legend.position = "bottom")
histo
ggsave("./Decontam/ITS_histograms.png",width = 15, height =30,
units = "in", dpi = 600 )
```

The clearest distinction can bee seen for the Prevelance method, so I will go with that one. However, Frequence 0.25 would also make sense, as would combined 0.25

```{r}
Check_contamination <- as.data.frame(sample_data(physeq_ITS_full_set)) # Put sample_data into a ggplot-friendly data.frame
Check_contamination$LibrarySize <- sample_sums(physeq_ITS_full_set)
Check_contamination <- Check_contamination[order(Check_contamination$LibrarySize),]
Check_contamination$Index <- seq(nrow(Check_contamination))
ggplot(data=Check_contamination, aes(x=Index, y=LibrarySize, color=sample_control)) + geom_point()
ggsave("./Decontam/Contamination_ITS_full.png")
```

Remove the contamination (I know, I could have just filtered on the p-value)

```{r}
contamdf.prev <- isContaminant(physeq_ITS_full_set, method="combined", conc=conc2, neg=neg2, threshold = 0.2, detailed=TRUE, normalize=TRUE)
table(contamdf.prev$contaminant)
```

328 contaminations were found, time to remove them

```{r}
# Make phyloseq object of presence-absence in negative controls and true samples
ps.pa.neg <- prune_samples(physeq_ITS_prev@sam_data$sample_control == "neg_control", physeq_ITS_prev)
ps.pa.pos <- prune_samples(sample_data(physeq_ITS_prev)$sample_control == "real_sample", physeq_ITS_prev)
# Make data.frame of prevalence in positive and negative samples
df.pa <- data.frame(pa.pos=taxa_sums(ps.pa.pos), pa.neg=taxa_sums(ps.pa.neg),
                      contaminant=contamdf.prev$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")
ps.noncontam <- prune_taxa(!contamdf.prev$contaminant, physeq_ITS_full_set)
ps.noncontam
```

```{r}
#save.image("./0.1.ITS_contamination.RData")
rm(list=setdiff(ls(), "physeq_bacterial_only"))
```

## 0. Filtering Phyloseq object

Filtering the metadata

```{r}
physeq_ITS <- ps.noncontam
metadata=read.csv("G:/My Drive/CityDeal/Metadata/Metadata_per_sample.csv", row.names=1)
samples_stoep <- rownames(metadata[(metadata$topology == "Stoep_bomen"),])
# In my case the metadata has multiple experiments in there. Here you could filter them on column
  metadata <- metadata[metadata$experiment=="CityDeal",]
  metadata <- metadata[!grepl("^Lin", rownames(metadata)),]
  metadata <- metadata[!grepl("^Gras", rownames(metadata)),]
  metadata <- subset(metadata, Sample_or_Control != "Control")
  #rownames(metadata)[rownames(metadata)=="T1A2B.C"] <- "T1A2B"
  metadata <- metadata[rownames(metadata) != "T1A2B.B", ]
  metadata <- metadata[rownames(metadata) != "T1A2B.A", ] 
# The filtering is now complete. Some of the metadata has to be translated though
metadata$topology <- gsub("Weg_bomen", "Street_with_trees", metadata$topology)
metadata$topology <- gsub("Stoep_bomen", "Sidewalk_with_trees", metadata$topology)
metadata$topology <- gsub("Stoep_geenbomen", "Sidewalk_without_trees", metadata$topology)
metadata$topology <- gsub("Weg_geenbomen", "Street_without_trees", metadata$topology)
metadata$topology <- gsub("Boomspiegel", "Treepit", metadata$topology)

## Add to phyloseq
META <- as.matrix(metadata)
META=sample_data(metadata)
sam_data(physeq_ITS) <- META
physeq_ITS
setdiff(rownames(metadata), rownames(physeq_ITS@sam_data))
# remove non-fungal
physeq_fungal_only <- subset_taxa(physeq_ITS, Kingdom == "Fungi")

physeq_fungal_only
table_nozeroes <- as.data.frame(physeq_fungal_only@otu_table) %>% filter(rowSums(as.data.frame(physeq_fungal_only@otu_table))>0)
dim(table_nozeroes)
mean(colSums(table_nozeroes))

setdiff(samples_stoep, rownames(sample_data(physeq_fungal_only)))

```

```{r}
save.image("./0.ITS_phyloseqs.RData")
rm(list=setdiff(ls(), "physeq_fungal_only"))
```

## 2. ITS Rarefaction

```{r}
sequencing_depth <- as.data.frame(sample_sums(physeq_fungal_only))
#write.csv(df, "G:/My Drive/CityDeal/Data_analysis_after_review/ITS/ITSsequencing_depth_raw.csv")
physeq_fungal_only_sub <- subset_samples(physeq_fungal_only, sample_sums(physeq_fungal_only)>=10000)
# rarefy your dataset to lowest sequencing length or a custom number, 711 should be kept constant:
phy_rare <- rarefy_even_depth(phy_rare,rngseed = 711, replace = F, trimOTUs = TRUE, 	verbose = F)
phy_rare
colSums(phy_rare@otu_table)
## We did lose 23 samples in this process. 
# Sample names before rarefaction
lost_samples <- setdiff(rownames(sample_data(physeq_fungal_only)),rownames(sample_data(phy_rare)))
lost_samples
colSums(otu_table(phy_rare))
```

### Supp fig 1; rarefaction curves

```{r}
## ITS rarefaction Curfe unrarefied
ITS_rarefaction_curve <- ggrare(physeq_fungal_only, step =500, color = "topology", se = F)

rarefaction_curve_ITS <- ITS_rarefaction_curve + mytheme + 
  scale_color_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECA1')) +
  labs( 
    x ="Number of reads",
    y ="Total ASVs"
    )
rarefaction_curve_ITS
ggsave("G:/My Drive/CityDeal/Figures/Rarefaction_ITS_raw_test.png",width = 15, height = 10, units = "in", dpi = 600)


## ITS rarefaction curve after rarefaction
ITS_rarefaction_curve_RF10000 <- ggrare(phy_rare, step = 100, color = "topology", se = F)

rarefaction_curve_ITS_RF <- ITS_rarefaction_curve_RF10000 + mytheme + 
  scale_color_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECA1')) +
  labs( 
    x ="Number of reads",
    y ="Total ASVs"
    )
rarefaction_curve_ITS_RF
ggsave("G:/My Drive/CityDeal/Figures/Rarefaction_ITS_RF11718_test.png",width = 15, height = 10, units = "in", dpi = 600)


```

```{r}
save.image("./2.ITS_Rarefaction.RData")
rm(ITS_rarefaction_curve,ITS_rarefaction_curve_RF10000,rarefaction_curve_ITS, rarefaction_curve_ITS_RF,sequencing_depth, lost_samples)
```

## 3. Fungal alpha

For statistics, I wrote an over-complicated function because I was relatively new at R and thought it would be more efficient. Ofcourse it was not, but I learned a lot in the process I guess. Since the function is there we might as well use it ;)

### Figure 2B

Testing alpha diversity after rarefaction

```{r}
## Rarefied; add richness parameters
Richnes_parameters <- estimate_richness(phy_rare, measures=c("Observed", "InvSimpson", "Shannon", "Chao1"))

enriched_meta <- merge(as.data.frame(phy_rare@sam_data), Richnes_parameters, by = "row.names", all.x = TRUE)
# And add groups (city-topology)
enriched_meta$group <- paste(enriched_meta$topology, enriched_meta$city)
write.csv(enriched_meta, "G:/My Drive/CityDeal/Data_analysis_after_review/ITS_enriched_meta_rarefied_test.csv")
phy_rare@sam_data <- sample_data(enriched_meta)

enriched_meta$topology = factor(enriched_meta$topology, levels = c("Treepit","Sidewalk_with_trees","Street_with_trees", "Sidewalk_without_trees", "Street_without_trees"))

enriched_meta = enriched_meta[order(enriched_meta$topology), ]
head(enriched_meta)
richness_all_c <- ggplot(data = enriched_meta,
                         aes(x=topology, y = Observed, fill=topology))+
  geom_boxplot(lwd=1.5)+
  #facet_grid(.~city, scales="free") + # For supplemental
  expand_limits(y=0:1000) +
  scale_y_continuous(breaks=seq(0,6000, by=200)) +
  labs(x="City", y="Richness ASVs", title="ASV Richness ITS")+
  #theme_bw()+
  scale_fill_manual(values=c('#FEECC8','#FFABAB','#DBFFD6', '#ACE7FF', '#ECD4FF')) +
  mytheme +
   theme(text = element_text(size=12),
         legend.text = element_text(size=12),
         axis.title=element_text(size=14),
         axis.text.x = element_text(size=12, angle = 90, hjust = 0.5, vjust=0.5),
         axis.text.y = element_text(size=12))
        #panel.grid.major = element_blank(), panel.grid.minor = element_blank())
richness_all_c


ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/ITS_Alpha_div_fig2.png", width = 15, height =10,units = "in", dpi = 600)

## For each city and topology apart
stats_rarefied_topo_city <- compare_variables_function(enriched_meta, "topology", "city", "Observed")
stats_rarefied_city_topo <- compare_variables_function(enriched_meta, "city", "topology", "Observed")
stats_rarefied_city_topo

#and for the total dataset
normality_rar <- shapiro.test(enriched_meta$Observed)
homogeinity_rar <- bartlett.test(enriched_meta$Observed, phy_rare@sam_data$group) 
# Check, we'll need Kruskall wallis

kruskal_rar <- kruskal.test(enriched_meta$Observed ~ enriched_meta$topology)
wilcox_rar <- pairwise.wilcox.test(enriched_meta$Observed, enriched_meta$topology, p.adjust.method = "BH")

wilcox_rar

kruskal_rar
metadata_rarefied <- enriched_meta
# metadata_rarefied <- as.data.frame(sample_data(phy_rare))  
# metadata_rarefied <- t(metadata_rarefied)
# metadata_rarefied <- as.data.frame(metadata_rarefied)
metadata_rarefied$Observed <- as.numeric(metadata_rarefied$Observed)
median_per_typology <- aggregate(Observed ~ topology, data = metadata_rarefied, median)

median_per_typology
```

### unrarefied sensitivity check

```{r}
## Rarefied; add richness parameters
Richnes_parameters <- estimate_richness(physeq_fungal_only, measures=c("Observed", "InvSimpson", "Shannon", "Chao1"))

enriched_meta <- merge(as.data.frame(physeq_fungal_only@sam_data), Richnes_parameters, by = "row.names", all.x = TRUE)
# And add groups (city-topology)
enriched_meta$group <- paste(enriched_meta$topology, enriched_meta$city)

physeq_fungal_only@sam_data <- sample_data(enriched_meta)

enriched_meta$topology = factor(enriched_meta$topology, levels = c("Treepit","Sidewalk_with_trees","Street_with_trees", "Sidewalk_without_trees", "Street_without_trees"))

enriched_meta = enriched_meta[order(enriched_meta$topology), ]
head(enriched_meta)
richness_all_c_unrar <- ggplot(data = enriched_meta,
                         aes(x=topology, y = Observed, fill=topology))+
  geom_boxplot(lwd=0.1)+
  #facet_grid(.~city, scales="free") +
  expand_limits(y=0:1000) +
  scale_y_continuous(breaks=seq(0,6000, by=1000)) +
  labs(x="City", y="Richness ASVs", title="ASV Richness ITS")+
  #theme_bw()+
  scale_fill_manual(values=c('#FEECC8','#FFABAB','#DBFFD6', '#ACE7FF', '#ECD4FF')) +
  mytheme +
   theme(text = element_text(size=12),
         legend.text = element_text(size=12),
         axis.title=element_text(size=14),
         axis.text.x = element_text(size=12, angle = 90, hjust = 0.5, vjust=0.5),
         axis.text.y = element_text(size=12))
        #panel.grid.major = element_blank(), panel.grid.minor = element_blank())
richness_all_c_unrar

ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/ITS_Alpha_div_unrar_supp.png", width = 15, height =10,
units = "in", dpi = 600)

## For each city and topology apart
stats_rarefied_topo_city_unrar <- compare_variables_function(enriched_meta, "topology", "city", "Observed")
stats_rarefied_city_topo_unrar <- compare_variables_function(enriched_meta, "city", "topology", "Observed")
stats_rarefied_city_topo_unrar

metadata_unrarefied <- enriched_meta

#and for the total dataset
normality_unrar <- shapiro.test(enriched_meta$Observed)
homogeinity_unrar <- bartlett.test(enriched_meta$Observed, physeq_fungal_only@sam_data$group) 
# Check, we'll need Kruskall wallis

kruskal_umrar <- kruskal.test(enriched_meta$Observed ~ enriched_meta$topology)
wilcox_umrar <- pairwise.wilcox.test(enriched_meta$Observed, enriched_meta$topology, p.adjust.method = "BH")

wilcox_umrar


```

Save some files

```{r}

metadata_rarefied <- as.data.frame(sample_data(phy_rare))

## Don't ask me why, but had to transform it twice, because it wouldn't save it otherwise. So I just accepted it and moved on with my life.
rownames(metadata_rarefied) <- metadata_rarefied$Row.names
metadata_rarefied <- t(metadata_rarefied)
write.csv(t(metadata_rarefied), "G:/My Drive/CityDeal/Data_analysis_after_review/ITS_metadata_rarefied_test.csv")


metadata_unrarefied <- as.data.frame(sample_data(physeq_fungal_only))
rownames(metadata_unrarefied) <- metadata_unrarefied$Row.names
metadata_unrarefied <- t(metadata_unrarefied)
write.csv(t(metadata_unrarefied), "G:/My Drive/CityDeal/Data_analysis_after_review/ITS_metadata_unrarefied_test.csv")

```

```{r}
save.image("./3.ITS_Alpha_diversity_Fig2.RData")
rm(list=setdiff(ls(), c("physeq_fungal_only_sub","phy_rare","mytheme")))
```

## 3.1 Taxonomomic analysis

```{r}
## Make presence absence phyloseq

## little fixup
#rownames(physeq_fungal_only@sam_data) <- physeq_fungal_only@sam_data$Row.names

physeq_fungal_only_pres_abs <- transform_sample_counts(physeq_fungal_only_sub, function(abund) 1*(abund>0))
physeq_fungal_only_pres_abs@sam_data$topology = factor(physeq_fungal_only_pres_abs@sam_data$topology, levels = c("Treepit","Sidewalk_with_trees","Street_with_trees", "Sidewalk_without_trees", "Street_without_trees"))
physeq_fungal_only_pres_abs@sam_data = physeq_fungal_only_pres_abs@sam_data[order(physeq_fungal_only_pres_abs@sam_data$topology), ]
physeq_fungal_only_pres_abs@sam_data
```

```{r}
barplot_ITS_taxa <- plot_grouped_taxa_bar(physeq_fungal_only_pres_abs, merge_var = "topology", level ="Class", higher_level = "Phylum", threshold = 3.5, palette = "Set3")
barplot_ITS_taxa

ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/ITS_Taxonomy_fig3.png", width = 15, height =10,units = "in", dpi = 600)

```

### Statistics

Ancom-bc is a good method to do the statistics behind this. It is compositional aware and does a (Sampling/pcr) bias correction. Input is normal phyloseq

```{r}
run_deseq2_phyloseq <- function(
  physeq,
  tax_rank = "Phylum",          # e.g. "Phylum", "Genus", or NULL for ASV
  design_formula = ~ city + topology,
  size_factor_type = "poscounts"
) {
topos <- levels(colData(dds_phylum)$topology)
pairs <- combn(topos, 2, simplify = FALSE)
  # 1. Aggregate if taxonomic rank is given
  if (!is.null(tax_rank)) {
    stopifnot(tax_rank %in% rank_names(physeq))
    physeq <- tax_glom(physeq, taxrank = tax_rank)
  }

  # 2. Remove zero-count taxa
  physeq <- prune_taxa(taxa_sums(physeq) > 0, physeq)

  # 3. Convert to DESeq2
  dds <- phyloseq_to_deseq2(physeq, design_formula)

  # 4. Estimate size factors (robust to zeros)
  dds <- estimateSizeFactors(dds, type = size_factor_type)

  # 5. Run DESeq2
  dds <- DESeq(dds)

  return(dds)
}

```

```{r}
dds_phylum <- run_deseq2_phyloseq(
  physeq_fungal_only,
  tax_rank = "Phylum",
  design_formula = ~ city + topology
)
levels(colData(dds_phylum)$topology)

all_contrasts <- lapply(pairs, function(p) {

  res <- results(
    dds_phylum,
    contrast = c("topology", p[1], p[2]),
    alpha = 0.05
  )

  as.data.frame(res) %>%
    rownames_to_column("Phylum") %>%
    mutate(
      contrast = paste(p[1], "vs", p[2])
    )
})

all_contrasts_df <- bind_rows(all_contrasts)

all_contrasts_df <- all_contrasts_df %>%
  mutate(padj_global = p.adjust(pvalue, method = "fdr"))
all_contrasts_df$ASV <- all_contrasts_df$Phylum
```

```{r}
## Atach taxonomy\
rep_ASVs <- all_contrasts_df$ASV
Phyla <- as.data.frame(physeq_bacterial_only@tax_table)
all_contrasts_df$Phylum <- Phyla[all_contrasts_df$ASV, "Phylum"]
write.csv(all_contrasts_df, "./3.1_ITS_Functional_analysis_statistics_order.csv")
```

Finding patterns:

```{r}

contrasts_sig_table <- all_contrasts_df %>%
  separate(contrast, into = c("g1", "g2"), sep = " vs ") %>%
  mutate(
    sig = padj_global < 0.05,
    direction = case_when(
      log2FoldChange > 0 ~ 1,
      log2FoldChange < 0 ~ -1,
      TRUE ~ 0
    )
  )


scores_contrasts <- contrasts_sig_table %>%
  mutate(
    g1_score = ifelse(sig, direction, 0),
    g2_score = ifelse(sig, -direction, 0)
  ) %>%
  select(Phylum, g1, g2, g1_score, g2_score) %>%
  pivot_longer(
    cols = c(g1, g2),
    names_to = "which",
    values_to = "group"
  ) %>%
  mutate(score = ifelse(which == "g1", g1_score, g2_score)) %>%
  group_by(Phylum, group) %>%
  summarise(score = sum(score), .groups = "drop")


nonsig_pairs <- contrasts_sig_table %>%
  filter(!sig) %>%
  select(Phylum, g1, g2)


build_pattern <- function(phylum) {
  
  s <- scores_contrasts%>% filter(Phylum == phylum) %>% arrange(desc(score))
  ns <- nonsig_pairs %>% filter(Phylum == phylum)

  groups <- list()
  
  for (g in s$group) {
    placed <- FALSE
    
    for (i in seq_along(groups)) {
      ref <- groups[[i]][1]
      
      tied <- ns %>%
        filter((g1 == g & g2 == ref) | (g1 == ref & g2 == g)) %>%
        nrow() > 0
      
      if (tied) {
        groups[[i]] <- c(groups[[i]], g)
        placed <- TRUE
        break
      }
    }
    
    if (!placed) groups[[length(groups) + 1]] <- g
  }
  
  paste(
    sapply(groups, function(x) paste(x, collapse = " = ")),
    collapse = " > "
  )
}


patterns_equal <- scores_contrasts %>%
  distinct(Phylum) %>%
  rowwise() %>%
  mutate(pattern = build_pattern(Phylum)) %>%
  ungroup()

filter(patterns_equal, Phylum == "Verrucomicrobiota")
```

```{r}
save.image("./3.1_ITS_Functional_analysis.RData")

```

## 4. PCOA - Beta Div

### CLR

We'll use a different kind of transformation for this:

```{r}

## Little patch-up
#sample_names(physeq_fungal_only)<- sample_data(physeq_fungal_only)$Row.names

## Perform CLR transformation
ps_clr_ITS <- transform_sample_counts(
  physeq_fungal_only_sub,
  function(x) clr(x + 1)
)

ps_clr_ITS
dist_aitch <- phyloseq::distance(ps_clr_ITS, method = "euclidean")
ord <- ordinate(ps_clr_ITS, method = "PCoA", distance = dist_aitch)
plot_ordination(ps_clr_ITS, ord, color = "topology", shape = "city")




# Add metadata
meta <- as.data.frame(sample_data(ps_clr_ITS))
meta$SampleID <- rownames(meta)
meta$Streetpair <- substr(meta$SampleID, 0, 3)
meta$Streetpair <- gsub("o","", meta$Streetpair)
meta$Streetpair <- substr(meta$Streetpair, 0, 2)
meta <- t(meta)
write.csv(t(meta), "G:/My Drive/CityDeal/Data_analysis_after_review/ITS_metadata_unrarefied.csv")

meta <- as.data.frame(t(meta))

# meta <- read.csv("./ITS_metadata_unrarefied_test.csv", row.names = 1)
# 
# # Extract sample scores
scores <- as.data.frame(ord$vectors)
scores$SampleID <- rownames(scores)
averages <- left_join(scores, meta, by = "SampleID")

df_sum <- averages %>%
  group_by(topology, city) %>%
  summarise(
    Axis1_mean = mean(Axis.1),
    Axis2_mean = mean(Axis.2),
    Axis3_mean = mean(Axis.3),
    Axis4_mean = mean(Axis.4),
    Axis1_se = sd(Axis.1) / sqrt(n()),
    Axis2_se = sd(Axis.2) / sqrt(n()),
    Axis3_se = sd(Axis.3) / sqrt(n()),
    Axis4_se = sd(Axis.4) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )


ggplot(df_sum, aes(Axis1_mean, Axis2_mean,
                   fill = topology, shape = city)) +
  geom_errorbar(
    aes(ymin = Axis2_mean - Axis2_se,
        ymax = Axis2_mean + Axis2_se),
    width = 0, 
    linewidth = 1
  ) +
  geom_errorbarh(
    aes(xmin = Axis1_mean - Axis1_se,
        xmax = Axis1_mean + Axis1_se),
    height = 0, 
    linewidth = 1
  ) +
  geom_point(size = 5, stroke = 1.5) +
  mytheme +
  scale_fill_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECC8')) +
  scale_shape_manual(values=c(21,22,23,24,25)) +
  labs(
    x = paste("PCoA1: ", round(ord$values$Relative_eig[1] * 100, 2), "%", sep=""),
    y=paste("PCoA2: ", round(ord$values$Relative_eig[2] * 100, 2), "%", sep="")) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  mytheme


ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/ITS_Eucledian_distance_PC1_PC2.png",width = 15, height = 10, units = "in", dpi = 600)

## Second and Third axis
ggplot(df_sum, aes(Axis2_mean, Axis3_mean,
                   fill = topology, shape = city)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = Axis3_mean - Axis3_se,
        ymax = Axis3_mean + Axis3_se),
    width = 0
  ) +
  geom_errorbarh(
    aes(xmin = Axis2_mean - Axis2_se,
        xmax = Axis2_mean + Axis2_se),
    height = 0
  ) +
  mytheme +
  scale_fill_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECC8')) +
  scale_shape_manual(values=c(21,22,23,24,25)) +
  labs(
    x = paste("PCoA2: ", round(ord$values$Relative_eig[2] * 100, 2), "%", sep=""),
    y=paste("PCoA3: ", round(ord$values$Relative_eig[3] * 100, 2), "%", sep="")) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  mytheme
ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/ITS_Eucledian_distance_PC2_PC3_test.png",width = 15, height = 10, units = "in", dpi = 600)

## Third and Fourth axis
ggplot(df_sum, aes(Axis3_mean, Axis4_mean,
                   fill = topology, shape = city)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = Axis4_mean - Axis4_se,
        ymax = Axis4_mean + Axis4_se),
    width = 0
  ) +
  geom_errorbarh(
    aes(xmin = Axis3_mean - Axis3_se,
        xmax = Axis3_mean + Axis3_se),
    height = 0
  ) +
  mytheme +
  scale_fill_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECC8')) +
  scale_shape_manual(values=c(21,22,23,24,25)) +
  labs(
    x = paste("PCoA3: ", round(ord$values$Relative_eig[3] * 100, 2), "%", sep=""),
    y=paste("PCoA4: ", round(ord$values$Relative_eig[4] * 100, 2), "%", sep="")) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  mytheme


ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/ITS_Eucledian_distance_PC3_PC4_test.png",width = 15, height = 10, units = "in", dpi = 600)
```

#### Permanova-clr

```{r}
adonis2(dist_aitch ~  topology, strata = meta$Streetpair,  data=as.data.frame(meta))
adonis2(dist_aitch ~  city, data=as.data.frame(meta))


beta_disp <- betadisper(dist_aitch, meta$topology)
anova(beta_disp)  # Tests if dispersion is significantly different
permutest(beta_disp)  # Permutation test for dispersion
permutest(beta_disp, pairwise = TRUE)


beta_disp <- betadisper(dist_aitch, meta$city)
anova(beta_disp)  # Tests if dispersion is significantly different
permutest(beta_disp)  # Permutation test for dispersion
permutest(beta_disp, pairwise = TRUE)
```

```{r}
## always difficulties with Phyloseq sample info to dataframe, fastest way is to just load the metadata csv. 
meta <- read.csv("G:/My Drive/CityDeal/Data_analysis_after_review/ITS_metadata_unrarefied.csv", row.names = 1)
# PERMANOVA for individual variables
adonis_city <- adonis2(dist_aitch ~ city, data=as.data.frame(meta))
adonis_city
adonis_topology <- adonis2(dist_aitch ~ topology, strata = meta$Streetpair, data= meta)
adonis_topology
adonis_streetname <- adonis2(dist_aitch ~ Streetpair, data= meta)
adonis_streetname
adonis_covering <- adonis2(dist_aitch ~ bedekking, data= meta)
adonis_covering
# adonis_typestraat <- adonis2(dist_aitch ~ type_straat, data= metadata_bact)
# adonis_covering
#adonis_covertree <- adonis2(dist_aitch ~ covertree, data = metadata_fungi)
#adonis_covertree
###Permanova assumes equal dispersion amongst groups
#beta_disp <- betadisper(dist_aitch, meta$Group)
#anova(beta_disp)  # Tests if dispersion is significantly different
#permutest(beta_disp)  # Permutation test for dispersion

# PERMANOVA for interaction (combination)
adonis_combination <- adonis2(dist_aitch ~ topology * city, data =meta)

PairPermITS <- pairwise.perm.manova(dist_aitch, meta$city, test="Pillai", nperm = 999, p.method = "fdr", F = TRUE, R2 = TRUE)
PairPermITS
PairPermITS$F.value

pairpermits2 <- pairwise.perm.manova(dist_aitch, meta$topology, test="Pillai", nperm = 999, p.method = "fdr", F = TRUE, R2 = TRUE)
pairpermits2
pairpermits2$F.value

street_pairperm <- pairwise.perm.manova(dist_aitch, meta$street, test = "Pillai", nperm =999, p.method = "fdr",F=T,R2=T)
# 
# pairperm_covertree <- pairwise.perm.manova(dist_aitch, metadata_fungi$covertree, test = "pillai", nperm =999, p.method ="fdr", F=T, R2=T)
# pairperm_covertree$R2.value

## For pairwase manova:
# higher F-value means that between-group variation is alrger relative to the within group variation
# This generally means a stronger effect of the grouping variable
# Display results

# Flatten and bind rows
combined_results <- bind_rows(
  data.frame(Test = "Adonis_topology", adonis_topology),
  data.frame(Test="Pairwise_permanova_topology", pairpermits2$p.value),
  data.frame(Test = "Pairwise_topology_R2", pairpermits2$R2.value),
  data.frame(Test = "Adonis_city", adonis_city),
  data.frame(Test = "Pairwise_permanova_city", PairPermITS$p.value),
  data.frame(Test = "R2_value_city", PairPermITS$R2.value),
  data.frame(Test = "Adonis_streetname", adonis_streetname),
  data.frame(Test = "Adonis_covering", adonis_covering),
  #data.frame(Test = "Adonis_typestreet", adonis_typestraat),
  #data.frame(Test = "Adonis_covertree", adonis_covertree),
  #data.frame(Test = "Pairwise_covertree", pairperm_covertree$p.value),
  #data.frame(Test = "R2_covertree", pairperm_covertree$R2.value),
  .id = "Source"
)
street_df <- bind_rows(
  data.frame(Test="p-value", street_pairperm$p.value),
  data.frame(Test="R2_values", street_pairperm$R2.value)
)

write.csv(combined_results, "G:/My Drive/CityDeal/Data_analysis_after_review/ITS_PCOA_statistical_summary_aitch_test.csv")

```

From the statistics I gather that the Streetname is the biggest determinant (so, samples from the same street are alike, geographic location). Therefore I think I should plot that...

I did not use this.

### Rarefied - Bray

```{r}
# Small fixups in data structure
meta <- read.csv("./ITS_metadata_rarefied.csv")
meta$Sample
 meta$X <- NULL
 rownames(meta) <- meta$Row.names

ITS_ordin_bray <- ordinate(phy_rare, method = "PCoA", distance = "bray")

df_bray <- merge(ITS_ordin_bray$vectors[,1:2], meta, by = "row.names")
row.names(df_bray) <- df_bray$Row.names
df_bray <- df_bray[, -1] 

# Extract sample scores
scores <- as.data.frame(ITS_ordin_bray$vectors)
scores$Row.names <- rownames(scores)
averages <- left_join(scores, meta, by = "Row.names")

df_sum <- averages %>%
  group_by(topology, city) %>%
  summarise(
    Axis1_mean = mean(Axis.1),
    Axis2_mean = mean(Axis.2),
    Axis3_mean = mean(Axis.3),
    Axis4_mean = mean(Axis.4),
    Axis1_se = sd(Axis.1) / sqrt(n()),
    Axis2_se = sd(Axis.2) / sqrt(n()),
    Axis3_se = sd(Axis.3) / sqrt(n()),
    Axis4_se = sd(Axis.4) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )


ggplot(df_sum, aes(Axis1_mean, Axis2_mean,
                   fill = topology, shape = city)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = Axis2_mean - Axis2_se,
        ymax = Axis2_mean + Axis2_se),
    width = 0
  ) +
  geom_errorbarh(
    aes(xmin = Axis1_mean - Axis1_se,
        xmax = Axis1_mean + Axis1_se),
    height = 0
  ) +
  mytheme +
  scale_fill_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECC8')) +
  scale_shape_manual(values=c(21,22,23,24,25)) +
  labs(
    x = paste("PCoA1: ", round(ITS_ordin_bray$values$Relative_eig[3] * 100, 2), "%", sep=""),
    y=paste("PCoA2: ", round(ITS_ordin_bray$values$Relative_eig[4] * 100, 2), "%", sep="")) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  mytheme

ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/Bray_ITS_rarefied_test.png",width = 15, height = 10, units = "in", dpi = 600)

pcoa_bray_all <- ggplot(df_sum, aes(Axis.1,Axis.2, fill=topology, shape=city)) +
  geom_point(size=5, alpha=0.7)+
  labs(x=paste("PCoA1: ", round(ITS_ordin_bray$values$Relative_eig[1] * 100, 2), "%", sep=""),
       y=paste("PCoA2: ", round(ITS_ordin_bray$values$Relative_eig[2] * 100, 2), "%", sep=""),
       title="Bray") +
  scale_fill_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECC8')) +
  scale_shape_manual(values=c(21,22,23,24,25)) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  mytheme

pcoa_bray_all

bray_distance <- phyloseq::distance(phy_rare, method = "bray")


adonis2(bray_distance ~ topology, strata = meta$Streetpair, data = meta, permutations = 1e4)
adonis2(bray_distance ~ topology, data = meta)


bd_topo <- betadisper(bray_distance, meta$topology,)
anova(bd_topo)
plot(bd_topo)
```

### Jaccard

```{r}
meta <- phy_rare@sam_data
rownames(meta) <- meta$Row.names

ITS_ordin_jaccard <- ordinate(phy_rare, method = "PCoA", distance = "jaccard")


df_jaccard <- merge(ITS_ordin_jaccard$vectors[,1:2], meta, by = "row.names")
row.names(df_jaccard) <- df_jaccard$Row.names
df_bray <- df_jaccard[, -1] 


pcoa_jaccard_all <- ggplot(df_jaccard, aes(Axis.1,Axis.2, fill=topology, shape=city)) +
  geom_point(size=5, alpha=0.7)+
  labs(x=paste("PCoA1: ", round(ITS_ordin_bray$values$Relative_eig[1] * 100, 2), "%", sep=""),
       y=paste("PCoA2: ", round(ITS_ordin_bray$values$Relative_eig[2] * 100, 2), "%", sep=""),
       title="jaccard ITS") +
  scale_fill_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECC8')) +
  scale_shape_manual(values=c(21,22,23,24,25)) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  mytheme 
  

pcoa_jaccard_all



ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/Jaccard_figure_please_delete.png",width = 15, height = 10, units = "in", dpi = 600)




jaccard_distance <- phyloseq::distance(phy_rare, method = "jaccard")

adonis2(jaccard_distance ~ topology, strata = meta$city, data = meta, permutations = 1e3)
adonis2(jaccard_distance ~ city, data = meta)


bd_topo <- betadisper(jaccard_distance, meta$topology,)
anova(bd_topo)
plot(bd_topo)
```

### Raw data - Bray

```{r}
meta <- read.csv("./ITS_metadata_unrarefied.csv", row.names = 1)


ITS_ordin_bray <- ordinate(physeq_fungal_only, method = "PCoA", distance = "bray")

df_bray <- merge(ITS_ordin_bray$vectors[,1:2], meta, by = "row.names")
row.names(df_bray) <- df_bray$Row.names
df_bray <- df_bray[, -1] 

# Extract sample scores
scores <- as.data.frame(ITS_ordin_bray$vectors)
scores$Row.names <- rownames(scores)
averages <- left_join(scores, meta, by = "Row.names")

df_sum <- averages %>%
  group_by(topology, city) %>%
  summarise(
    Axis1_mean = mean(Axis.1),
    Axis2_mean = mean(Axis.2),
    Axis3_mean = mean(Axis.3),
    Axis4_mean = mean(Axis.4),
    Axis1_se = sd(Axis.1) / sqrt(n()),
    Axis2_se = sd(Axis.2) / sqrt(n()),
    Axis3_se = sd(Axis.3) / sqrt(n()),
    Axis4_se = sd(Axis.4) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )


ggplot(df_sum, aes(Axis1_mean, Axis2_mean,
                   fill = topology, shape = city)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = Axis2_mean - Axis2_se,
        ymax = Axis2_mean + Axis2_se),
    width = 0
  ) +
  geom_errorbarh(
    aes(xmin = Axis1_mean - Axis1_se,
        xmax = Axis1_mean + Axis1_se),
    height = 0
  ) +
  mytheme +
  scale_fill_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECC8')) +
  scale_shape_manual(values=c(21,22,23,24,25)) +
  labs(
    x = paste("PCoA1: ", round(ord$values$Relative_eig[3] * 100, 2), "%", sep=""),
    y=paste("PCoA2: ", round(ord$values$Relative_eig[4] * 100, 2), "%", sep="")) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  mytheme

bray_distance <- phyloseq::distance(physeq_fungal_only, method = "bray")

adonis2(bray_distance ~ type_straat, strata = meta$city, data = meta, permutations = 1e3)
adonis2(bray_distance ~ city, data = meta)

```

### Conclusion ITS - Alpha/beta

Every tested matrix shows clearly that city is a defining factor for beta-diversity, while topology does not matter at all. Combined with the alpha diversity this means that the same fungi are found everywhere, but the treepit has a way higher number of fungi per sample. Baas Becking.

```{r}
save.image("./4.PCoA_ITS_CLR.RData")
rm(list=setdiff(ls(), c("physeq_fungal_only_sub","phy_rare","ps_clr_ITS","mytheme")))
```

# 16S

## 5. Making phyloseq object

```{r}
#reading files, you need four inputs: 
#otu table, taxonomy table, metadata, and rooted tree

ASV_table_16S =read.csv("./Input_data/Full_ASV_table_16s.csv", row.names=1, header = T)
view(head(ASV_table_16S))
#rownames(ASV_table_16S) <- ASV_table_16S$X
#otu_table=as.matrix(ASV_table_16S)
new_colnames <- colnames(ASV_table_16S)
split_colnames <- lapply(new_colnames, str_split_i,pattern= "_", i=1)
split_colnames <- unlist(split_colnames)
colnames(ASV_table_16S) <- split_colnames

taxonomy=read.csv("G:/My Drive/CityDeal/16S - Panji/16s_taxonomy.csv", row.names=1, header = T)
replace(taxonomy, taxonomy == " ", NA) -> tax1
taxonomy1=as.matrix(tax1)

Metadata_FA_16S <- read.csv("./Input_data/FA_metadata_16S.csv")



 
# metadata=read.csv("G:/My Drive/CityDeal/16S - Panji/metadata.csv", row.names=1)
# metadata
#META=sample_data(metadata)
rownames(Metadata_FA_16S) <- Metadata_FA_16S$SampleID

META<-sample_data(Metadata_FA_16S)
OTU=otu_table(ASV_table_16S, taxa_are_rows=TRUE)
TAX=tax_table(taxonomy1)



#new_sample_names <- as.list(make.unique(sapply(strsplit(sample_names(OTU),"_"), `[`, 1)))
#sample_names(OTU) <- new_sample_names

#make the phyloseq object for downstream analysis
physeq_16S_full=phyloseq(OTU, TAX, META)


physeq_16S_full



############ Filter OTU table ############


```

clean data

```{r}
save.image("./5.1.S16_phyloseq_full")
rm(list=setdiff(ls(), "physeq_16S_full"))
```

## 6. Decontamination

```{r}
#Little patch up
# physeq_16S@sam_data$Sample_or_Control <- physeq_16S@sam_data$sample
# physeq_16S@sam_data$sample <- NULL

Check_contamination <- as.data.frame(sample_data(physeq_16S_full)) # Put sample_data into a ggplot-friendly data.frame
Check_contamination$LibrarySize <- sample_sums(physeq_16S_full)
Check_contamination <- Check_contamination[order(Check_contamination$LibrarySize),]
Check_contamination$Index <- seq(nrow(Check_contamination))
ggplot(data=Check_contamination, aes(x=Index, y=LibrarySize, color=sample_control)) + geom_point()
```

Oke, there is clearly some major contamination in two plates

```{r}
#settings
theme_set(theme_bw())
options(stringsAsFactors = FALSE)
path.in <- "G:/My Drive/CityDeal/Data_analysis_after_review/Input_data" # CHANGE ME
path.out <- "G:/My Drive/CityDeal/Data_analysis_after_review/Decontam" # CHANGE ME

#color palettes
cbbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7","#000000")
methodPalette <- c(Frequency="darkorange2", Prevalence="turquoise3", Combined="darkorchid4")
prevalencePalette <- c("2" = "#edf8e9", "3-5" = "#bae4b3", "6-10" = "#74c476", "11+" = "#238b45")

```

Check decontamination

```{r}
#physeq_16S_full@sam_data$ng.uL <- as.numeric(physeq_16S_full@sam_data$ng.uL)
physeq_16S_no_control <- subset_samples(physeq_16S_full, sample_control == "real_sample")
conc2 <- physeq_16S_no_control@sam_data$ng.uL # ng/ul
neg2 <- physeq_16S_full@sam_data$sample_control == 'neg_control' # True if a negative control sample
# Frequency-based contaminant classification
ocf <- isContaminant(physeq_16S_no_control, conc=conc2, threshold=0.2, detailed=TRUE, normalize=TRUE, method='frequency')
# Prevalence-based contaminant classification
ocp <- isContaminant(physeq_16S_full, neg=neg2, threshold=0.2, detailed=TRUE, normalize=TRUE, method='prevalence')
# Combined contaminant classification
occ <- isContaminant(physeq_16S_full, conc=conc2, neg=neg2, threshold=0.2, detailed=TRUE, normalize=TRUE, method='combined')
```

figures

```{r}

probcols <- data.frame(row.names=rownames(ocf),prob.f=ocf$p.freq, prob.p=ocp$p.prev, prob.c = occ$p)

TAX2 <- as.matrix(physeq_16S_full@tax_table)
TAX2 <- cbind(TAX2, probcols[rownames(TAX2),])
### MANY ASVs have NA, these all occur in only one sample. I will not remove them, because frankly we can be quite sure they are no contaminants
# Crude comparison of frequency and prevalence contaminant assignment
table(probcols$prob.f<0.1, probcols$prob.c<0.1)
# Augment with prevalence groups and Abundance column
physeq_16S_prev <- transform_sample_counts(physeq_16S_full, function(x) ifelse(x>0, 1, 0))
TAX2$TAXprev <- taxa_sums(physeq_16S_prev)[rownames(TAX2)] # Add prevalences
TAX2$Prevalence <- cut(TAX2$TAXprev, c(0, 2, 5, 10, 9999), labels=c("2", "3-5", "6-10", "11+"))

TAXann <- rbind(cbind(TAX2, Score=TAX2$prob.f, Method="Frequency"),
                cbind(TAX2, Score=TAX2$prob.p, Method="Prevalence"),
                cbind(TAX2, Score=TAX2$prob.c, Method="Combined"))

TAXann$Method <- factor(TAXann$Method, levels=c('Frequency','Prevalence','Combined'))
prevalencePalette <- c("2" = "#edf8e9", "3-5" = "#bae4b3", "6-10" = "#74c476", "11+" = "#238b45")



TAXannTest <- subset(TAXann, Method != "Prevalence")
#figure 2b
histo <- ggplot(TAXann, aes(x=Score, fill=Prevalence))

histo <- histo + geom_histogram() + labs(x = 'decontam Score', y='Number ASVs') + 
  facet_wrap(~Method, nrow=1) +
  #http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=4
  scale_fill_manual(values=prevalencePalette) +
  scale_y_sqrt()+
 # theme(axis.text.x = element_text(colour="grey20",size=10,angle=0,
 #                                   hjust=.5,vjust=.5,face="plain"),
 #       axis.text.y = element_text(colour="grey20",size=10,angle=0,
 #                                  hjust=1,vjust=0,face="plain"),
 #       axis.title.x = element_text(colour="grey20",size=15,angle=0,
 #                                   hjust=.5,vjust=0,face="plain"),
 #       axis.title.y = element_text(colour="grey20",size=15,angle=90,
 #                                   hjust=.5,vjust=.5,face="plain"),
 #       strip.text.x = element_text(size = 15, angle = 0),
 #       legend.title = element_text(size = 15, angle = 0),
 #       legend.text = element_text(size = 10, angle = 0)) +
  theme(legend.position = "bottom")
histo
ggsave("./Decontam/16S_histograms.png",width = 15, height =30,
units = "in", dpi = 600 )
```

```{r}
# Make phyloseq object of presence-absence in negative controls and true samples
ps.pa <- transform_sample_counts(physeq_16S_full, function(abund) 1*(abund>0))
ps.pa.neg <- prune_samples(ps.pa@sam_data$sample_control == "neg_control", ps.pa)
ps.pa.pos <- prune_samples(sample_data(ps.pa)$sample_control == "real_sample", ps.pa)
# Make data.frame of prevalence in positive and negative samples
df.pa <- data.frame(pa.pos=taxa_sums(ps.pa.pos), pa.neg=taxa_sums(ps.pa.neg),
                      contaminant=contamdf.prev$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")
ps.noncontam <- prune_taxa(!occ$contaminant, physeq_16S_full)
ps.noncontam
```

```{r}
#Remove Mitochondria, Chloroplast, and unnecessary taxons 

physeq_bacterial_only <- subset_taxa(ps.noncontam, Kingdom == "Bacteria")
physeq_bacterial_only <- prune_samples(sample_data(physeq_bacterial_only)$sample_control == "real_sample", physeq_bacterial_only)
physeq_bacterial_only


```

```{r}
#save.image("./6.S16_contamination.RData")
rm(list=setdiff(ls(), "physeq_bacterial_only"))
```

## 6.1 subsetting phyloseq

```{r}
metadata=read.csv("G:/My Drive/CityDeal/Metadata/Metadata_per_sample.csv", row.names=1)
# In my case the metadata has multiple experiments in there. Here you could filter them on column
  metadata <- metadata[metadata$experiment=="CityDeal",]
  metadata <- metadata[!grepl("^Lin", rownames(metadata)),]
  metadata <- metadata[!grepl("^Gras", rownames(metadata)),]
  metadata <- subset(metadata, Sample_or_Control != "Control")
 #metadata <- metadata[rownames(metadata) != "T1A2B.C", ]<- "T1A2B"
  metadata <- metadata[rownames(metadata) != "T1A2B.B", ]
  metadata <- metadata[rownames(metadata) != "T1A2B.A",] 
# The filtering is now complete. Some of the metadata has to be translated though
metadata$topology <- gsub("Weg_bomen", "Street_with_trees", metadata$topology)
metadata$topology <- gsub("Stoep_bomen", "Sidewalk_with_trees", metadata$topology)
metadata$topology <- gsub("Stoep_geenbomen", "Sidewalk_without_trees", metadata$topology)
metadata$topology <- gsub("Weg_geenbomen", "Street_without_trees", metadata$topology)
metadata$topology <- gsub("Boomspiegel", "Treepit", metadata$topology)

## Add to phyloseq
META <- as.matrix(metadata)
META=sample_data(metadata)
sam_data(physeq_bacterial_only) <- META
rownames(physeq_bacterial_only@sam_data["T1A2B.C",]) <- "T1A2B"
physeq_bacterial_only
```

```{r}
save.image("./6.1.S16_phyloseq_citydeal")
rm(list=setdiff(ls(), "physeq_bacterial_only"))
```

## 7. Rarefaction 16S

```{r}
sequencing_depth <- as.data.frame(sample_sums(physeq_bacterial_only))



write.csv(sequencing_depth, "G:/My Drive/CityDeal/Data_analysis_after_review/16Ssequencing_depth_raw.csv")
# rarefy your dataset to lowest sequencing length or a custom number, 711 should be kept constant:
phy_rare <- subset_samples(physeq_bacterial_only, sample_sums(physeq_bacterial_only)>=40000)
phy_rare <- rarefy_even_depth(phy_rare,rngseed = 711, replace = F, trimOTUs = TRUE, 	verbose = F)
phy_rare
colSums(phy_rare@otu_table)


# Sample names before rarefaction
lost_samples <- setdiff(rownames(sample_data(physeq_bacterial_only)),rownames(sample_data(phy_rare)))
lost_samples
colSums(otu_table(phy_rare))

```

```{r}
#save.image("./7.S16_Rarefaction.RData")
rm(list=setdiff(ls(), c("physeq_bacterial_only", "phy_rare")))
```

### Supplemental figure

```{r}
## ITS rarefaction Curfe unrarefied
rarefaction_curve_16S <- ggrare(physeq_bacterial_only, step =2000, color = "topology", se = F)

rarefaction_curve_16S <- rarefaction_curve_16S + mytheme + 
  scale_color_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECA1')) +
  labs( 
    x ="Number of reads",
    y ="Total ASVs"
    )
rarefaction_curve_16S
ggsave("G:/My Drive/CityDeal/Figures/Rarefaction_16S_raw.png",width = 15, height = 10, units = "in", dpi = 600)


## ITS rarefaction curve after rarefaction
rarefaction_curve_16S_RF44000 <- ggrare(phy_rare, step = 1000, color = "topology", se = F)

rarefaction_curve_16S_RF44000 <- rarefaction_curve_16S_RF44000 + mytheme + 
  scale_color_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECA1')) +
  labs( 
    x ="Number of reads",
    y ="Total ASVs"
    )
rarefaction_curve_16S_RF44000
ggsave("G:/My Drive/CityDeal/Figures/Rarefaction_16S_RF42363.png.png",width = 15, height = 10, units = "in", dpi = 600)

```

```{r}
save.image("./7.S16_Rarefaction.RData")
rm(list=setdiff(ls(), c("physeq_bacterial_only","phy_rare","mytheme")))
```

## 8. Bacterial alpha

```{r}
## Rarefied; add richness parameters
Richnes_parameters_rar <- estimate_richness(phy_rare, measures=c("Observed", "InvSimpson", "Shannon", "Chao1"))

enriched_meta <- merge(as.data.frame(phy_rare@sam_data), Richnes_parameters_rar, by = "row.names", all.x = TRUE)
# And add groups (city-topology)
enriched_meta$group <- paste(enriched_meta$topology, enriched_meta$city)
write.csv(enriched_meta, "G:/My Drive/CityDeal/Data_analysis_after_review/16S_enriched_meta_rarefied.csv")
#enriched_meta <- read.csv("G:/My Drive/CityDeal/Data_analysis_after_review/16S_enriched_meta_rarefied.csv", row.names = 1)

phy_rare@sam_data <- sample_data(enriched_meta)

enriched_meta$topology = factor(enriched_meta$topology, levels = c("Tree pit","sidewalk with trees","street with trees", "sidewalk without trees", "street without trees"))

enriched_meta = enriched_meta[order(enriched_meta$topology), ]
head(enriched_meta)
richness_all_c <- ggplot(data = enriched_meta,
                         aes(x=topology, y = Observed, fill=topology))+
  geom_boxplot(lwd=1.5)+
  #facet_grid(.~city, scales="free") + # For supplemental
  expand_limits(y=0:1000) +
  scale_y_continuous(breaks=seq(0,9000, by=2000)) +
  labs(x="City", y="Richness ASVs", title="ASV Richness 16S rarefied")+
  #theme_bw()+
  scale_fill_manual(values=c('#FEECC8','#FFABAB','#DBFFD6', '#ACE7FF', '#ECD4FF')) +
  mytheme +
   theme(text = element_text(size=12),
         legend.text = element_text(size=12),
         axis.title=element_text(size=14),
         axis.text.x = element_text(size=12, angle = 90, hjust = 0.5, vjust=0.5),
         axis.text.y = element_text(size=12))
        #panel.grid.major = element_blank(), panel.grid.minor = element_blank())
richness_all_c


ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/16S_alpha_Div_figure_2.png", width = 15, height =10,
units = "in", dpi = 600)

## For each city and topology apart
stats_rarefied_topo_city <- compare_variables_function(enriched_meta, "topology", "city", "Observed")
stats_rarefied_city_topo <- compare_variables_function(enriched_meta, "city", "topology", "Observed")
stats_rarefied_topo_city

#and for the total dataset
normality_rar <- shapiro.test(enriched_meta$Observed)
homogeinity_rar <- bartlett.test(enriched_meta$Observed, phy_rare@sam_data$group) 
# Check, we'll need Kruskall wallis

kruskal_rar <- kruskal.test(enriched_meta$Observed ~ enriched_meta$topology)
wilcox_rar <- pairwise.wilcox.test(enriched_meta$Observed, enriched_meta$topology, p.adjust.method = "BH")

kruskal_rar
wilcox_rar
```

There are no differences in alpha diversity between cities and/or typologies, Amserdam is slightly higher than Dordrecht

### unrarefied sensitivity check

```{r}
##Little fix
#sample_names(physeq_bacterial_only) <- sample_data(physeq_bacterial_only)$Row.names

## Rarefied; add richness parameters
Richnes_parameters <- estimate_richness(physeq_bacterial_only, measures=c("Observed", "InvSimpson", "Shannon", "Chao1"))



enriched_meta <- merge(as.data.frame(physeq_bacterial_only@sam_data), Richnes_parameters, by = "row.names", all.x = TRUE)
rownames(enriched_meta) <- enriched_meta$Row.names
# And add groups (city-topology)
enriched_meta$group <- paste(enriched_meta$topology, enriched_meta$city)

#write.csv(t(enriched_meta), "./16S_metadata_unrarefied.csv")
enriched_meta <- read.csv("./16S_metadata_unrarefied.csv")


enriched_meta$topology = factor(enriched_meta$topology, levels = c("Tree pit","sidewalk with trees","street with trees", "sidewalk without trees", "street without trees"))

enriched_meta = enriched_meta[order(enriched_meta$topology), ]
head(enriched_meta)
richness_all_c_unrar <- ggplot(data = enriched_meta,
                         aes(x=topology, y = Observed, fill=topology))+
  geom_boxplot(lwd=0.1)+
  #facet_grid(.~city, scales="free") +
  expand_limits(y=0:1000) +
  scale_y_continuous(breaks=seq(0,6000, by=1000)) +
  labs(x="City", y="Richness ASVs", title="ASV Richness ITS")+
  #theme_bw()+
  scale_fill_manual(values=c('#FEECC8','#FFABAB','#DBFFD6', '#ACE7FF', '#ECD4FF')) +
  mytheme +
   theme(text = element_text(size=12),
         legend.text = element_text(size=12),
         axis.title=element_text(size=14),
         axis.text.x = element_text(size=12, angle = 90, hjust = 0.5, vjust=0.5),
         axis.text.y = element_text(size=12))
        #panel.grid.major = element_blank(), panel.grid.minor = element_blank())
richness_all_c_unrar

#ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/16S_Alpha_div_unrar_supp.png", width = 15, height =10,
units = "in", dpi = 600)

## For each city and topology apart
stats_rarefied_topo_city_unrar <- compare_variables_function(enriched_meta, "topology", "city", "Observed")
stats_rarefied_city_topo_unrar <- compare_variables_function(enriched_meta, "city", "topology", "Observed")
stats_rarefied_city_topo_unrar

#and for the total dataset
normality_unrar <- shapiro.test(enriched_meta$Observed)
homogeinity_unrar <- bartlett.test(enriched_meta$Observed, physeq_bacterial_only@sam_data$group) 
# Check, we'll need Kruskall wallis

kruskal_umrar <- kruskal.test(enriched_meta$Observed ~ enriched_meta$topology)
wilcox_umrar <- pairwise.wilcox.test(enriched_meta$Observed, enriched_meta$topology, p.adjust.method = "BH")

kruskal_umrar
wilcox_umrar
```

Unrarefied shows not patterns either.

```{r}
#save.image("./8.S16_Alpha_diversity.RData")
rm(list=setdiff(ls(), c("physeq_bacterial_only","phy_rare","mytheme")))
```

## 8.1 Taxonomic analysis

```{r}
## Make presence absence phyloseq

## little fixup
#rownames(physeq_fungal_only@sam_data) <- physeq_fungal_only@sam_data$Row.names

physeq_bacterial_only_pres_abs <- transform_sample_counts(physeq_bacterial_only, function(abund) 1*(abund>0))
physeq_bacterial_only_pres_abs@sam_data$topology = factor(physeq_bacterial_only_pres_abs@sam_data$topology, levels = c("Tree pit","sidewalk with trees","street with trees", "sidewalk without trees", "street without trees"))
physeq_bacterial_only_pres_abs@sam_data = physeq_bacterial_only_pres_abs@sam_data[order(physeq_bacterial_only_pres_abs@sam_data$topology), ]
physeq_bacterial_only_pres_abs@sam_data
```

```{r}
barplot_16S_taxa <- plot_grouped_taxa_bar(physeq_bacterial_only_pres_abs, merge_var = "topology", level ="Class", higher_level = "Phylum", threshold = 3.5, palette = "Set3")
barplot_16S_taxa

barplot_16S_rare <- plot_rare_phyla_bar(physeq_bacterial_only_pres_abs, merge_var = "topology",  higher_level = "Phylum", threshold = 3.5, palette = "Set3")
barplot_16S_rare

ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/S16_Taxonomy_fig3_low_orders.png", width = 15, height =10,units = "in", dpi = 600)

```

```{r}
#save.image("./3.1_ITS_Functional_analysis.RData")

```

### Statistics

Ancom-bc is a good method to do the statistics behind this. It is compositional aware and does a (Sampling/pcr) bias correction. Input is normal phyloseq

```{r}
run_deseq2_phyloseq <- function(
  physeq,
  tax_rank = "Order",          # e.g. "Phylum", "Genus", or NULL for ASV
  design_formula = ~ city + topology,
  size_factor_type = "poscounts"
) {

  # 1. Aggregate if taxonomic rank is given
  if (!is.null(tax_rank)) {
    stopifnot(tax_rank %in% rank_names(physeq))
    physeq <- tax_glom(physeq, taxrank = tax_rank)
  }

  # 2. Remove zero-count taxa
  physeq <- prune_taxa(taxa_sums(physeq) > 0, physeq)

  # 3. Convert to DESeq2
  dds <- phyloseq_to_deseq2(physeq, design_formula)

  # 4. Estimate size factors (robust to zeros)
  dds <- estimateSizeFactors(dds, type = size_factor_type)

  # 5. Run DESeq2
  dds <- DESeq(dds)

  return(dds)
}

```

```{r}
dds_phylum <- run_deseq2_phyloseq(
  physeq_bacterial_only,
  tax_rank = "Phylum",
  design_formula = ~ city + topology
)
levels(colData(dds_phylum)$topology)
topos <- levels(colData(dds_phylum)$topology)
pairs <- combn(topos, 2, simplify = FALSE)
all_contrasts <- lapply(pairs, function(p) {

  res <- results(
    dds_phylum,
    contrast = c("topology", p[1], p[2]),
    alpha = 0.05
  )

  as.data.frame(res) %>%
    rownames_to_column("Phylum") %>%
    mutate(
      contrast = paste(p[1], "vs", p[2])
    )
})

all_contrasts_df <- bind_rows(all_contrasts)

all_contrasts_df <- all_contrasts_df %>%
  mutate(padj_global = p.adjust(pvalue, method = "fdr"))
all_contrasts_df$ASV <- all_contrasts_df$Phylum
```

```{r}
## Atach taxonomy\
rep_ASVs <- all_contrasts_df$ASV
Phyla <- as.data.frame(physeq_bacterial_only@tax_table)
all_contrasts_df$Order <- Phyla[all_contrasts_df$ASV, "Phylum"]
write.csv(all_contrasts_df, "./8.1_16S_taxonomy_analysis_statistics_phylum.csv")
```

fj

```{r}
contrasts_sig_table <- all_contrasts_df %>%
  separate(contrast, into = c("g1", "g2"), sep = " vs ") %>%
  mutate(
    sig = padj_global < 0.05,
    direction = case_when(
      log2FoldChange > 0 ~ 1,
      log2FoldChange < 0 ~ -1,
      TRUE ~ 0
    )
  )


scores_contrasts <- contrasts_sig_table %>%
  mutate(
    g1_score = ifelse(sig, direction, 0),
    g2_score = ifelse(sig, -direction, 0)
  ) %>%
  select(Phylum, g1, g2, g1_score, g2_score) %>%
  pivot_longer(
    cols = c(g1, g2),
    names_to = "which",
    values_to = "group"
  ) %>%
  mutate(score = ifelse(which == "g1", g1_score, g2_score)) %>%
  group_by(Phylum, group) %>%
  summarise(score = sum(score), .groups = "drop")


nonsig_pairs <- contrasts_sig_table %>%
  filter(!sig) %>%
  select(Phylum, g1, g2)


build_pattern <- function(phylum) {
  
  s <- scores_contrasts%>% filter(Phylum == phylum) %>% arrange(desc(score))
  ns <- nonsig_pairs %>% filter(Phylum == phylum)

  groups <- list()
  
  for (g in s$group) {
    placed <- FALSE
    
    for (i in seq_along(groups)) {
      ref <- groups[[i]][1]
      
      tied <- ns %>%
        filter((g1 == g & g2 == ref) | (g1 == ref & g2 == g)) %>%
        nrow() > 0
      
      if (tied) {
        groups[[i]] <- c(groups[[i]], g)
        placed <- TRUE
        break
      }
    }
    
    if (!placed) groups[[length(groups) + 1]] <- g
  }
  
  paste(
    sapply(groups, function(x) paste(x, collapse = " = ")),
    collapse = " > "
  )
}


patterns_equal <- scores_contrasts %>%
  distinct(Phylum) %>%
  rowwise() %>%
  mutate(pattern = build_pattern(Phylum)) %>%
  ungroup()

write.csv(patterns_equal, "./16S_taxonomy_statistical_summary_patterns.csv")
```

```{r}
save.image("./8.1.S16_Taxonomy.RData")
```

New visualisation

```{r}
ps_norm <- transform_sample_counts(
  ps_counts,
  function(x) x / phyloseq::sample_sums(ps_counts)
)
```

## 9. PCoA - Beta Div

### CLR

```{r}

## Little patch-up
#sample_names(physeq_bacterial_only)<- sample_data(physeq_bacterial_only)$Row.names

## Perform CLR transformation
ps_clr_S16 <- transform(physeq_bacterial_only, "clr")

ps_clr_S16

dist_aitch <- distance(ps_clr_S16, method = "euclidean")
ord <- ordinate(ps_clr_S16, method = "PCoA", distance = "euclidean")
str(ord)
plot_ordination(ps_clr_S16, ord, color = "city", shape = "topology")

## TRY
clr_mat_16S <- t(as(otu_table(ps_clr_S16), "matrix"))

pca_16S<- prcomp(clr_mat_16S, scale. = FALSE)

plot(pca_16S$x[,1], pca_16S$x[,2])

df_16S <- data.frame(
  pca_16S$x[,1:2],
  sample_data(ps_clr_S16)
)

library(ggplot2)
ggplot(df_16S, aes(PC1, PC2, color = city, shape = topology)) +
  geom_point(size = 3)


# Add metadata
 meta <- as.data.frame(sample_data(ps_clr_S16))
 meta$SampleID <- rownames(meta)
 meta$Streetpair <- substr(meta$SampleID, 0, 3)
 meta$Streetpair <- gsub("o","", meta$Streetpair)
 meta$Streetpair <- substr(meta$Streetpair, 0, 2)
# meta <- t(meta)
#  write.csv(t(meta), "G:/My Drive/CityDeal/Data_analysis_after_review/S16_metadata_unrarefied.csv")
meta <- as.data.frame(meta)

# Extract sample scores
scores <- as.data.frame(ord$vectors)
scores$SampleID <- rownames(scores)
averages <- left_join(scores, meta, by = "SampleID")

df_sum <- averages %>%
  group_by(topology, city) %>%
  summarise(
    Axis1_mean = mean(Axis.1),
    Axis2_mean = mean(Axis.2),
    Axis3_mean = mean(Axis.3),
    Axis4_mean = mean(Axis.4),
    Axis1_se = sd(Axis.1) / sqrt(n()),
    Axis2_se = sd(Axis.2) / sqrt(n()),
    Axis3_se = sd(Axis.3) / sqrt(n()),
    Axis4_se = sd(Axis.4) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )


ggplot(df_sum, aes(Axis1_mean, Axis2_mean,
                   fill = topology, shape = city)) +
  geom_errorbar(
    aes(ymin = Axis2_mean - Axis2_se,
        ymax = Axis2_mean + Axis2_se),
    width = 0,
    linewidth = 1
  ) +
  geom_errorbarh(
    aes(xmin = Axis1_mean - Axis1_se,
        xmax = Axis1_mean + Axis1_se),
    height = 0,
    linewidth = 1
  ) +
    geom_point(size = 5, stroke = 1.5) +
  mytheme +
  scale_fill_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECC8')) +
  scale_shape_manual(values=c(21,22,23,24,25)) +
  labs(
    x = paste("PCoA1: ", round(ord$values$Relative_eig[1] * 100, 2), "%", sep=""),
    y=paste("PCoA2: ", round(ord$values$Relative_eig[2] * 100, 2), "%", sep="")) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  mytheme


ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/S16_Eucledian_distance_PC1_PC2.png",width = 15, height = 10, units = "in", dpi = 600)

## Second and Third axis
ggplot(df_sum, aes(Axis2_mean, Axis3_mean,
                   fill = topology, shape = city)) +
  geom_errorbar(
    aes(ymin = Axis3_mean - Axis3_se,
        ymax = Axis3_mean + Axis3_se),
    width = 0
  ) +
  geom_errorbarh(
    aes(xmin = Axis2_mean - Axis2_se,
        xmax = Axis2_mean + Axis2_se),
    height = 0,
    linewidth =1 
  ) +
  geom_point(size = 5, stroke = 1.5) +
  mytheme +
  scale_fill_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECC8')) +
  scale_shape_manual(values=c(21,22,23,24,25)) +
  labs(
    x = paste("PCoA2: ", round(ord$values$Relative_eig[2] * 100, 2), "%", sep=""),
    y=paste("PCoA3: ", round(ord$values$Relative_eig[3] * 100, 2), "%", sep="")) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  mytheme
ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/S16_Eucledian_distance_PC2_PC3.png",width = 15, height = 10, units = "in", dpi = 600)

## Third and Fourth axis
ggplot(df_sum, aes(Axis3_mean, Axis4_mean,
                   fill = topology, shape = city)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = Axis4_mean - Axis4_se,
        ymax = Axis4_mean + Axis4_se),
    width = 0
  ) +
  geom_errorbarh(
    aes(xmin = Axis3_mean - Axis3_se,
        xmax = Axis3_mean + Axis3_se),
    height = 0
  ) +
  mytheme +
  scale_fill_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECC8')) +
  scale_shape_manual(values=c(21,22,23,24,25)) +
  labs(
    x = paste("PCoA3: ", round(ord$values$Relative_eig[3] * 100, 2), "%", sep=""),
    y=paste("PCoA4: ", round(ord$values$Relative_eig[4] * 100, 2), "%", sep="")) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  mytheme


ggsave("G:/My Drive/CityDeal/Data_analysis_after_review/S16_Eucledian_distance_PC3_PC4.png",width = 15, height = 10, units= "in", dpi = 600)
```

```{r}
## Stupid fix
# meta <- t(meta)
# write.csv(t(meta), "./16S_metadata_unrarefied.csv")
meta <- read.csv("./16S_metadata_unrarefied.csv", row.names = 1)

perm_typology_pcity <- adonis2(as.dist(dist_aitch) ~  city, data=meta )
perm_typology_pcity
bd_topo <- betadisper(dist_aitch, meta$city)
aov(bd_topo)

```

Statistics by clr

```{r}
# always difficulties with Phyloseq sample info to dataframe, fastest way is to just load the metadata csv. 
#meta <- read.csv("G:/My Drive/CityDeal/Data_analysis_after_review/S16_metadata_unrarefied.csv", row.names = 1)
# PERMANOVA for individual variables
adonis_city <- adonis2(dist_aitch ~ city, data=as.data.frame(meta))
adonis_city
adonis_topology <- adonis2(dist_aitch ~ topology, strata = meta$Streetpair, data= meta)
adonis_topology
adonis_streetname <- adonis2(dist_aitch ~ streetname, data= meta)
adonis_streetname
adonis_covering <- adonis2(dist_aitch ~ covering, data= meta)
adonis_covering
# adonis_typestraat <- adonis2(dist_aitch ~ type_straat, data= metadata_bact)
# adonis_covering
#adonis_covertree <- adonis2(dist_aitch ~ covertree, data = metadata_fungi)
#adonis_covertree
##Permanova assumes equal dispersion amongst groups
beta_disp <- betadisper(dist_aitch, meta$topology)
anova(beta_disp)  # Tests if dispersion is significantly different
permutest(beta_disp)  # Permutation test for dispersion

# PERMANOVA for interaction (combination)
adonis_combination <- adonis2(dist_aitch ~ topology * city, data =meta)

PairPermS16 <- pairwise.perm.manova(dist_aitch, meta$city, test="Pillai", nperm = 999, p.method = "fdr", F = TRUE, R2 = TRUE)
PairPermS16
PairPermS16$F.value

pairpermits2 <- pairwise.perm.manova(dist_aitch, meta$topology, test="Pillai", nperm = 999, p.method = "fdr", F = TRUE, R2 = TRUE)
pairpermits2
pairpermits2$F.value

street_pairperm <- pairwise.perm.manova(dist_aitch, meta$street, test = "Pillai", nperm =999, p.method = "fdr",F=T,R2=T)
# 
# pairperm_covertree <- pairwise.perm.manova(dist_aitch, metadata_fungi$covertree, test = "pillai", nperm =999, p.method ="fdr", F=T, R2=T)
# pairperm_covertree$R2.value

## For pairwase manova:
# higher F-value means that between-group variation is alrger relative to the within group variation
# This generally means a stronger effect of the grouping variable
# Display results

# Flatten and bind rows
combined_results <- bind_rows(
  data.frame(Test = "Adonis_topology", adonis_topology),
  data.frame(Test="Pairwise_permanova_topology", pairpermits2$p.value),
  data.frame(Test = "Pairwise_topology_R2", pairpermits2$R2.value),
  data.frame(Test = "Adonis_city", adonis_city),
  data.frame(Test = "Pairwise_permanova_city", PairPermS16$p.value),
  data.frame(Test = "R2_value_city", PairPermS16$R2.value),
  data.frame(Test = "Adonis_streetname", adonis_streetname),
  data.frame(Test = "Adonis_covering", adonis_covering),
  #data.frame(Test = "Adonis_typestreet", adonis_typestraat),
  #data.frame(Test = "Adonis_covertree", adonis_covertree),
  #data.frame(Test = "Pairwise_covertree", pairperm_covertree$p.value),
  #data.frame(Test = "R2_covertree", pairperm_covertree$R2.value),
  .id = "Source"
)
street_df <- bind_rows(
  data.frame(Test="p-value", street_pairperm$p.value),
  data.frame(Test="R2_values", street_pairperm$R2.value)
)

write.csv(combined_results, "G:/My Drive/CityDeal/Data_analysis_after_review/S16_PCOA_statistical_summary_aitch.csv")
```

```{r}
save.image("./9.S16-BetaDiversiy.RData")

```

```{r}
beta_disp <- betadisper(dist_aitch, meta$city)
anova(beta_disp)  # Tests if dispersion is significantly different
permutest(beta_disp)  # Permutation test for dispersion
permutest(beta_disp, pairwise = TRUE)

```

### Rarefied - Bray

```{r}
# Small fixups in data structure
meta <- read.csv("./16S_metadata_rarefied.csv")
meta$Sample
 meta$X <- NULL
 rownames(meta) <- meta$Row.names

S16_ordin_bray <- ordinate(phy_rare, method = "PCoA", distance = "bray")

df_bray <- merge(S16_ordin_bray$vectors[,1:2], meta, by = "row.names")
row.names(df_bray) <- df_bray$Row.names
df_bray <- df_bray[, -1] 

# Extract sample scores
scores <- as.data.frame(S16_ordin_bray$vectors)
scores$Row.names <- rownames(scores)
averages <- left_join(scores, meta, by = "Row.names")

df_sum <- averages %>%
  group_by(topology, city) %>%
  summarise(
    Axis1_mean = mean(Axis.1),
    Axis2_mean = mean(Axis.2),
    Axis3_mean = mean(Axis.3),
    Axis4_mean = mean(Axis.4),
    Axis1_se = sd(Axis.1) / sqrt(n()),
    Axis2_se = sd(Axis.2) / sqrt(n()),
    Axis3_se = sd(Axis.3) / sqrt(n()),
    Axis4_se = sd(Axis.4) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )


pcoa_bray_all<-ggplot(df_sum, aes(Axis1_mean, Axis2_mean,
                   fill = topology, shape = city)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = Axis2_mean - Axis2_se,
        ymax = Axis2_mean + Axis2_se),
    width = 0
  ) +
  geom_errorbarh(
    aes(xmin = Axis1_mean - Axis1_se,
        xmax = Axis1_mean + Axis1_se),
    height = 0
  ) +
  mytheme +
  scale_fill_manual(values=c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECC8')) +
  scale_shape_manual(values=c(21,22,23,24,25)) +
  labs(
    x = paste("PCoA1: ", round(ord$values$Relative_eig[3] * 100, 2), "%", sep=""),
    y=paste("PCoA2: ", round(ord$values$Relative_eig[4] * 100, 2), "%", sep="")) +
  guides(fill=guide_legend(override.aes=list(shape=21)))+
  mytheme

pcoa_bray_all

ggsave("./S16_bray.png",width = 15, height = 10, units = "in", dpi = 600)




bray_distance <- phyloseq::distance(phy_rare, method = "bray")


adonis2(bray_distance ~ city + topology, strata = meta$street, data = meta, by = "margin")
pairwise.perm.manova(bray_distance, meta$Plate_number, test="Pillai", nperm = 999, p.method = "fdr", F = TRUE, R2 = TRUE)
adonis2(bray_distance ~ city, strata = meta$topology, data = meta)


bd_topo <- betadisper(bray_distance, meta$topology,)
anova(bd_topo)
plot(bd_topo)
```

# Functional heatmaps

### 16s

```{r}
library(microeco)
library(file2meco)
library(vegan)
library(microbiome)

# Loading the FAPROTAX database is more complex. To calculate annotation, we first transform the phyloseq to the meco object
Meco_16S_Rare   <- phyloseq2meco(physeq_filtered_16s)
Meco_16S_Func   <- trans_func$new(dataset =Meco_16S_Rare)


# Calculate the FAPROTAX 1/0 matrix and extraction 
Meco_16S_Func$cal_spe_func(prok_database = "FAPROTAX")
Func16S         <- as.data.frame(Meco_16S_Func$res_spe_func)
sum(rowSums(Func16S, na.rm = F) == 0)
sum(rowSums(Func16S, na.rm = F) != 0)
sum(rowSums(Func16S, na.rm = F) != 0)/length(rownames(Func16S))
## 26688 ASV are annotated which is 17%

# Note that I will continue with this microeco package, as translating this back into phyloseq is not ideal
```

#### Bar Graph/Heat Maps

```{r}
# Calculating percentage coverage and making a heatmap
Meco_16S_Func$cal_spe_func_perc(abundance_weighted = TRUE, perc = TRUE, dec = 8)
Func16SPerc     <- as.data.frame(Meco_16S_Func$res_spe_func_perc)

# creating heatmap x axis order from Ordination (same as phyloseq heatmap)
## extracting ordination coordinates and calculating radial theta
Ord16STest      <- Meco_16S_Rare$clone()
Ord16STest$cal_betadiv(method = "jaccard")
Ord16SBeta_group      <- trans_beta$new(dataset = Ord16STest, group = "group", measure = "jaccard")
Ord16SBeta$cal_ordination(method = "PCoA")
Scores          <- scores(Ord16SBeta$res_ordination, choices = c(1,2), display = "sites")
Order16S        <- Ord16STest$sample_names()[order(radial_theta(Scores))]

rm(Ord16STest, Ord16SBeta, Scores)

# Creating heatmap
Meco_16S_Func$plot_spe_func_perc(add_facet = TRUE,
                                 order_x = Order16S)+
    scale_fill_gradient2(low = "#FF0000", mid ="#CCFFFF", high = "#001A1A" ,midpoint = 0.00001, na.value = "#FF0000")+
    labs(title = "Heatmap of functional diversity (16S data)")


ggsave(paste(datafolder,"Functional_analysis_ASV/16S_Func_Heat.png"), device = png, width = 2000, height = 1200, units = "px", dpi = 144, limitsize = FALSE)
ggsave(paste(datafolder,"Functional_analysis_ASV/ASV_16S_Func_Heat2.png"), device = png, width = 6000, height = 1200, units = "px", dpi = 144, limitsize = FALSE)
#Note the one sample in the middle that is different is B077; Higher levels of chloroplasts, photoautotrophy and also  human associated: maybe Biofilm? Algae? Moss?

# Merged By topology
Meco_16S_Funcx  <- Meco_16S_Rare$merge_samples("topology")

Meco_16S_Funcx  <- trans_func$new(Meco_16S_Funcx)
Meco_16S_Funcx$cal_spe_func(prok_database = "FAPROTAX")
Func16Sx         <- as.data.frame(Meco_16S_Funcx$res_spe_func)

Meco_16S_Funcx$cal_spe_func_perc(abundance_weighted = TRUE, perc = FALSE, dec = 8)
Func16SxPerc     <- as.data.frame(Meco_16S_Funcx$res_spe_func_perc)

# creating order from chaos
Ord16STest      <- Meco_16S_Rare$clone()
Ord16STest      <- Ord16STest$merge_samples("topology")
Ord16STest$cal_betadiv(method = "jaccard")
Ord16SBeta      <- trans_beta$new(dataset = Ord16STest, measure = "jaccard")
Ord16SBeta$cal_ordination(method = "PCoA")
Scores          <- scores(Ord16SBeta$res_ordination, choices = c(1,2), display = "sites")
Order16SGroup   <- Ord16STest$sample_names()[order(radial_theta(Scores))]

rm(Ord16STest, Ord16SBeta, Scores)

# plot

Meco_16S_Funcx$plot_spe_func_perc(add_facet = TRUE,
                                 order_x = Order16SGroup) +
    scale_fill_gradient2(low = "#FF0000", mid = "#CCFFFF", high = "#001A1A",midpoint = 0.00001, na.value = "#FF0000")+
    labs(title = "Heatmap of functional diversity grouped by topology (16S data)")

write.csv(Func16SxPerc, paste(datafolder,"Functional_analysis_ASV/ASV_16S_Func_topology.csv")
ggsave(paste(datafolder,"Functional_analysis_ASV/ASV_16S_Func_topology.png"), device = png, width = 2000, height = 1200, units = "px", dpi = 144, limitsize = FALSE)

#Meco_16S_Rare$sample_table$group <- paste(Meco_16S_Rare$sample_table$city, #Meco_16S_Rare$sample_table$topology, sep = "_")

# Merged By group
Meco_16S_Funcx  <- Meco_16S_Rare$merge_samples("topology")

Meco_16S_Funcx  <- trans_func$new(Meco_16S_Funcx)
Meco_16S_Funcx$cal_spe_func(prok_database = "FAPROTAX")
Func16Sx         <- as.data.frame(Meco_16S_Funcx$res_spe_func)

Meco_16S_Funcx$cal_spe_func_perc(abundance_weighted = TRUE, perc = FALSE, dec = 8)
Func16SxPerc     <- as.data.frame(Meco_16S_Funcx$res_spe_func_perc)

# creating order from chaos
Ord16STest      <- Meco_16S_Rare$clone()
Ord16STest      <- Ord16STest$merge_samples("topology")
Ord16STest$cal_betadiv(method = "jaccard")
Ord16SBeta      <- trans_beta$new(dataset = Ord16STest, measure = "jaccard")
Ord16SBeta$cal_ordination(method = "PCoA")
Scores<- scores(Ord16SBeta$res_ordination, choices = c(1,2), display = "sites")
Order16SGroup   <- Ord16STest$sample_names()[order(radial_theta(Scores))]

rm(Ord16STest, Ord16SBeta, Scores)

library(viridis)
library(scales)
# plot
Meco_16S_Funcx$res_spe_func_perc_trans$value <- ifelse(Meco_16S_Funcx$res_spe_func_perc_trans$value==NA,0,Meco_16S_Funcx$res_spe_func_perc_trans$value)

Meco_16S_Funcx$plot_spe_func_perc(add_facet = TRUE,
                                 order_x = Order16SGroup) +
    scale_fill_viridis_d(option = "viridis", na.value = "white")+
    labs(title = "Heatmap of functional diversity grouped by city and topology (16S data)")

Meco_16S_Funcx$plot_spe_func_perc(add_facet = TRUE,
                                  order_x = Order16SGroup) +
  scale_fill_viridis(
    option = "viridis",
    values = rescale(c(0, 0.002, 0.01, 0.3)),  # more detail in the low range
    na.value = "white",
    direction = -1                             # optional: reverse if needed
  ) +
  labs(title = "Heatmap of functional diversity grouped by city and topology (16S data)")

bold_labels <- function(labels, to_bold) {
  sapply(labels, function(label) {
    if (label %in% to_bold) {
      label <- paste0("*", label)
      bquote(bold(.(label)))
    } else {
      label
    }
  })
}
rows_to_make_bold <- as.list(names(signif_functions))

Meco_16S_Funcx$plot_spe_func_perc(add_facet = TRUE,
                                  order_x = c("sidewalk with trees", "sidewalk without trees", "street with trees", "street without trees", "Tree pit")) +
  scale_fill_viridis(
  option = "viridis",
  values = rescale(c(0, 0.002, 0.01, 0.3)),
  breaks = c(0, 0.005, 0.01, 0.05, 0.1, 0.3),
  labels = scales::label_percent(accuracy = 0.1),
  na.value = "white",
  direction = -1,
  guide = guide_colorbar(barheight = 40, barwidth = 1)
)+
  scale_y_discrete(labels = bold_labels(colnames(Meco_16S_Funcx$res_spe_func), rows_to_make_bold)) +  
  labs(title = "Heatmap of functional diversity grouped by city and topology (16S data)")




  #write.csv(Func16SxPerc, paste(datafolder,"Functional_analysis_ASV/ASV_16S_Func_City_and_topology.csv")
ggsave(paste(datafolder,"Functional_analysis_ASV/ASV_16S_Func_city_and_topology_viridis_test.png"), device = png, width = 2000, height = 2400, units = "px", dpi = 144, limitsize = FALSE)
```

#### Statistics

```{r}

library(rstatix)
## We make a dataframe with metadata and functional data combined.
func_table_with_meta <- cbind(Func16SPerc, as.data.frame(physeq_filtered_16s@sam_data))
write.csv(func_table_with_meta ,paste(datafolder,"/Functional_analysis_ASV/test_functional_table_merged.csv")

# Run Kruskal-Wallis test for each function, based on topology. KW because not all values are normally distributed and we should not combine tests.
kw_results <- apply(func_table_with_meta[1:ncol(Func16SPerc)], 2, function(x) kruskal.test(x ~ func_table_with_meta$topology)$p.value)

# Adjust p-values for multiple testing (Benjamini-Hochberg)
kw_results_adj <- p.adjust(kw_results, method = "BH")

# View top significant functions
signif_functions <- sort(kw_results_adj[kw_results_adj < 0.05])
names(signif_functions)
str(names(signif_functions))
func_table_signif <- func_table[,names(signif_functions)]



# Practise to do all pairwise comps.
# for (i in 1:length(names(func_table_signif))){
#   name <- names(signif_functions)[i]
#   pairwise_result <- pairwise_wilcox_test(
#     data.frame(value = func_table_signif[,i], group = category),
#     value ~ group,
#     p.adjust.method = "BH"
#     )
#   pairwise_result_list[name]<- pairwise_result
# }
#   pairwise_result



func_table_signif_merge <-  merge(func_table_signif,physeq_filtered_16s@sam_data,by="row.names",all.x=TRUE)
rownames(func_table_signif_merge) <- func_table_signif_merge$Row.names
func_table$Row.names <- NULL
#One keeps giving an error, so I decided to just remove it.
#names(signif_functions[56])
results_list <- list()
for (func in colnames(func_table_signif)) {
  print(func)
  # Perform pairwise test
  test <- func_table_signif_merge %>%
    dunn_test(as.formula(paste(func, "~ topology")), p.adjust.method = "BH") %>%
    add_significance()
  
  # Add median direction info
  median_vals <- func_table_signif_merge %>%
    group_by(topology) %>%
    summarise(med = median(.data[[func]], na.rm = TRUE))
  
  test <- test %>%
    left_join(median_vals, by = c("group1" = "topology")) %>%
    rename(med1 = med) %>%
    left_join(median_vals, by = c("group2" = "topology")) %>%
    rename(med2 = med) %>%
    mutate(direction = case_when(
      p.adj < 0.05 & med1 > med2 ~ paste(group1, ">", group2),
      p.adj < 0.05 & med1 < med2 ~ paste(group1, "<", group2),
      TRUE ~ "NS"
    )) %>%
    mutate(Function = func)
  
  results_list[[func]] <- test
}

# Combine all test results
all_results <- bind_rows(results_list)

write.csv(all_results, paste(datafolder,"/Functional_analysis_ASV/pairwise_comparison_functions_topology.csv")

```

### ITS

```{r}
# Having switched to Microeco for functional trait analysis, most of the ITS process has been commented out.

# # FungalTraits
# # extracting the taxtable from phyloseq (rarefied)
# TAXITS <- as.data.frame(tax_table(ASV_ITS_Rare))
# TAXITS1 <- TAXITS
# 
# # Adding guild annotation
# TAXITS1$Genus <- gsub("g__", "", TAXITS1$Genus)
# TAXITS1$Guild <- as.factor(Fungaltraits$guild_fg[match(TAXITS1$Genus, Fungaltraits$Genus)])
# TAXITS1 <- TAXITS1[,c(8,1,2,3,4,5,6,7)]
# # guild has 36 levels
# 
# #new phyloseq
# TAXITS1 <- tax_table(as.matrix(TAXITS1))
# ASV_ITS_Func1 <- phyloseq(otu_table(ASV_ITS_Rare),TAXITS1,Treetment)
# rm(TAXITS1)

library(file2meco)
library(microeco)
#Microeco. Note that these have a more recent database
Meco_ITS_Rare   <- phyloseq2meco(physeq_fungal_only)
Meco_ITS_Func   <- trans_func$new(Meco_ITS_Rare)
Meco_ITS_Func$cal_spe_func(fungi_database = "FungalTraits")
FuncITS         <- as.data.frame(Meco_ITS_Func$res_spe_func)
sum(rowSums(FuncITS, na.rm = F) == 0)
sum(rowSums(FuncITS, na.rm = F) != 0)
sum(rowSums(FuncITS, na.rm = F) != 0)/length(rownames(FuncITS))
# Coverage has improved to 37% (at 14.000 annotations)
```

#### Heath_map

```{r}
# FungalTraits x Microeco
# Calculating percentage coverage and making a heatmap
Meco_ITS_Func$cal_spe_func_perc(abundance_weighted = TRUE, perc = FALSE, dec = 8)
FuncITSPerc     <- as.data.frame(Meco_ITS_Func$res_spe_func_perc)

## extracting ordination coordinates and calculating radial theta
OrdITSTest      <- Meco_ITS_Rare$clone()
OrdITSTest$cal_betadiv(method = "bray")
OrdITSBeta      <- trans_beta$new(dataset = OrdITSTest, group = "topology", measure = "bray")
OrdITSBeta$cal_ordination(method = "PCoA")
Scores          <- scores(OrdITSBeta$res_ordination, choices = c(1,2), display = "sites")
OrderITS        <- OrdITSTest$sample_names()[order(radial_theta(Scores))]

rm(OrdITSTest, OrdITSBeta, Scores)

Meco_ITS_Func$plot_spe_func_perc(add_facet = TRUE,
                                 order_x = OrderITS)+
    scale_fill_gradient2(mid = "#FFCC99", high = "#1A0D00", na.value = "#FF0000")+
    labs(title = "Heatmap of functional diversity (ITS data)")
ggsave("ASV_ITS_Func_Heat_test.png", device = png, width = 2000, height = 1500, units = "px", dpi = 144, limitsize = FALSE)
ggsave("ASV_ITS_Func_Heat2.png", device = png, width = 6000, height = 1500, units = "px", dpi = 144, limitsize = FALSE)
# note that B077 on the right 10th percentile has very low fungal species richness and low functional percentages

# Merged By city and topology
Meco_ITS_Funcx  <- Meco_ITS_Rare$merge_samples("topology")
Meco_ITS_Funcx  <- trans_func$new(Meco_ITS_Funcx)

Meco_ITS_Funcx$cal_spe_func(fungi_database = "FungalTraits")
FuncITSx         <- as.data.frame(Meco_ITS_Funcx$res_spe_func)

Meco_ITS_Funcx$cal_spe_func_perc(abundance_weighted = TRUE, perc = FALSE, dec = 8)
FuncITSxPerc     <- as.data.frame(Meco_ITS_Funcx$res_spe_func_perc)

# Order in the court
OrdITSTest      <- Meco_ITS_Rare$clone()
OrdITSTest      <- OrdITSTest$merge_samples("topology")
OrdITSTest$cal_betadiv(method = "bray")
OrdITSBeta      <- trans_beta$new(dataset = OrdITSTest, measure = "bray")
OrdITSBeta$cal_ordination(method = "PCoA")
Scores          <- scores(OrdITSBeta$res_ordination, choices = c(1,2), display = "sites")
OrderITSGroup   <- OrdITSTest$sample_names()[order(radial_theta(Scores))]

rm(OrdITSTest, OrdITSBeta, Scores)

Meco_ITS_Funcx$plot_spe_func_perc(add_facet = TRUE,
                                 order_x = OrderITSGroup) +
    scale_fill_gradient2(mid = "#FFCC99", high = "#1A0D00", na.value = "#FF0000")+
    labs(title = "Heatmap of functional diversity grouped by substrate (ITS data)")


ggsave("ASV_ITS_Func_city_and_topology_test.png", device = png, width = 2000, height = 1500, units = "px", dpi = 144, limitsize = FALSE)


# Merged By topology
Meco_ITS_Funcx  <- Meco_ITS_Rare$merge_samples("topology")
Meco_ITS_Funcx  <- trans_func$new(Meco_ITS_Funcx)

Meco_ITS_Funcx$cal_spe_func(fungi_database = "FungalTraits")
FuncITSx         <- as.data.frame(Meco_ITS_Funcx$res_spe_func)

Meco_ITS_Funcx$cal_spe_func_perc(abundance_weighted = TRUE, perc = FALSE, dec = 8)
FuncITSxPerc     <- as.data.frame(Meco_ITS_Funcx$res_spe_func_perc)

# Order in the court
OrdITSTest      <- Meco_ITS_Rare$clone()
OrdITSTest      <- OrdITSTest$merge_samples("topology")

OrdITSTest$cal_betadiv(method = "jaccard")
OrdITSBeta      <- trans_beta$new(dataset = OrdITSTest, measure = "jaccard")
OrdITSBeta$cal_ordination(method = "PCoA")
Scores          <- scores(OrdITSBeta$res_ordination, choices = c(1,2), display = "sites")
OrderITSGroup   <- OrdITSTest$sample_names()[order(radial_theta(Scores))]

rm(OrdITSTest, OrdITSBeta, Scores)


```

#### Statistics

```{r}
# Assuming your function table is `func_table` and metadata is `meta_table`
# Ensure rows in func_table match sample order in meta_table
func_table <- FuncITSPerc[rownames(physeq_fungal_only@sam_data), ]
category <- as.factor(physeq_fungal_only@sam_data$city)

# Run Kruskal-Wallis test for each function
kw_results <- apply(func_table, 2, function(x) kruskal.test(x ~ category)$p.value)

# Adjust p-values for multiple testing (e.g., Benjamini-Hochberg)
kw_results_adj <- p.adjust(kw_results, method = "BH")

# View top significant functions
signif_functions <- sort(kw_results_adj[kw_results_adj < 0.05])
names(signif_functions)
str(names(signif_functions))
func_table_signif <- func_table[,names(signif_functions)]

library(rstatix)

func_table_signif_merge <-  merge(func_table_signif,phy_rare@sam_data,by="row.names",all.x=TRUE)
rownames(func_table_signif_merge) <- func_table_signif_merge$Row.names
func_table$Row.names <- NULL
names(func_table_signif_merge) <- make.names(names(func_table_signif_merge))
names(func_table_signif) <- make.names(names(func_table_signif))
results_list <- list()
for (func in colnames(func_table_signif)) {
  print(func)
  # Perform pairwise test
  test <- func_table_signif_merge %>%
    dunn_test(as.formula(paste(func, "~ topology")), p.adjust.method = "BH") %>%
    add_significance()
  
  # Add median direction info
  median_vals <- func_table_signif_merge %>%
    group_by(topology) %>%
    summarise(med = median(.data[[func]], na.rm = TRUE))
  
  test <- test %>%
    left_join(median_vals, by = c("group1" = "topology")) %>%
    rename(med1 = med) %>%
    left_join(median_vals, by = c("group2" = "topology")) %>%
    rename(med2 = med) %>%
    mutate(direction = case_when(
      p.adj < 0.05 & med1 > med2 ~ paste(group1, ">", group2),
      p.adj < 0.05 & med1 < med2 ~ paste(group1, "<", group2),
      TRUE ~ "NS"
    )) %>%
    mutate(Function = func)
  
  results_list[[func]] <- test
}

# Combine all test results
all_results <- bind_rows(results_list)

write.csv(all_results,"ITS_pairwise_comparison_functions.csv")

```

Run after stats

```{r}
# Meco_ITS_Funcx$plot_spe_func_perc(add_facet = TRUE,
#                                  order_x = OrderITSGroup) +
#     scale_fill_gradient2(mid = "#FFCC99", high = "#1A0D00", na.value = "#FF0010")+
#     labs(title = "Heatmap of functional diversity grouped by substrate (ITS data)")
bold_labels <- function(labels, to_bold) {
  sapply(labels, function(label) {
    if (label %in% to_bold) {
      label <- paste0("*", label)
      bquote(bold(.(label)))
    } else {
      bquote(.(label))
    }
  }, simplify = FALSE)
}

rows_to_make_bold <- sub(".*\\|", "", names(signif_functions))
intersect(colnames(Meco_ITS_Funcx$res_spe_func), rows_to_make_bold)
clean_labels <- sub(".*\\|", "", colnames(Meco_ITS_Funcx$res_spe_func))
sig_funcs <- intersect(rows_to_make_bold, colnames(Meco_ITS_Funcx$res_spe_func))

sig_matrix <- Meco_ITS_Funcx$res_spe_func[, sig_funcs, drop = FALSE]

tmp <- Meco_ITS_Funcx
tmp$res_spe_func <- sig_matrix


tmp$plot_spe_func_perc(add_facet = TRUE,
                       order_x = c("Treepit", "Sidewalk_with_trees", "Street_with_trees", "Sidewalk_without_trees", "Street_without_trees")) +
  scale_fill_viridis(
    option = "viridis",
    values = rescale(c(0, 0.1, 0.4, 0.8)),
    labels = scales::label_percent(accuracy = 0.5),
    na.value = "white",
    direction = 1,
    guide = guide_colorbar(barheight = 40, barwidth = 1)
  ) +
  labs(title = "Heatmap of significant functional diversity (ITS data)") +
  theme(
    axis.text.y = element_text(size = 14)   # increase function label size
  )

ggsave("ASV_ITS_Functopology_viridis_significant_only.png", device = png, width = 15, height = 26, units = "in", dpi = 144, limitsize = FALSE)
```

```{r}
rows_to_keep <- unique(rows_to_make_bold)

tmp <- Meco_ITS_Funcx
tmp$res_spe_func <- Meco_ITS_Funcx$res_spe_func[, 
                     colnames(Meco_ITS_Funcx$res_spe_func) %in% rows_to_keep]

tmp$plot_spe_func_perc(add_facet = TRUE,
                       order_x = OrderITSGroup) +
  scale_fill_viridis(
    option = "viridis",
    values = rescale(c(0, 0.1, 0.3, 0.8)),
    labels = scales::label_percent(accuracy = 0.1),
    na.value = "white",
    direction = 1,
    guide = guide_colorbar(barheight = 40, barwidth = 1)
  ) +
  labs(title = "Heatmap of significant functional diversity grouped by topology (ITS data)") +
  theme(axis.text.y = element_text(size = 14))

ggsave("ASV_ITS_Functopology_viridis_significant_only_test.png", device = png, width = 15, height = 26, units = "in", dpi = 144, limitsize = FALSE)
```
