---
title: "Multicolinearity detection"
output: html_document
date: "2024-11-08"
editor_options: "Sjoerd"
  chunk_output_type: inline
---

```{r}
mytheme <- theme_bw()+
  theme(text = element_text(size=15),
        strip.background = element_blank(),
        strip.placement = "outside",
        legend.box.background = element_rect(),
        legend.box.margin = margin(1, 1, 1, 1),
        legend.title = element_text(face = "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) 
```

## Basic statistics

```{r}
# Load tidyverse for data manipulation
library(tidyverse)
library(broom)
library(openxlsx)
# --- 1. Read and clean the data ---

# Read CSV
Soil_parameters <- read.csv("G:/My Drive/CityDeal/Eurofins_results/Soil_analysis_for_statistics.csv", row.names = 1, sep = ";", dec = ".", na.strings = c("", " ", "<", "<0.5", "<1", "<1.1", "<0.1", "<200", "<150"))

# Ensure Group is treated as a factor
Soil_parameters$Group <- as.factor(Soil_parameters$Group)

# Replace "<" patterns by NA and convert numeric columns properly
soil_par_clean <- Soil_parameters %>%
  mutate(across(where(is.character), ~ gsub("<.*", NA, .))) %>%   # remove '<' values
  mutate(across(where(is.character), ~ as.numeric(.)))             # convert to numeric when possible

# --- 2. Compute group means for all numeric columns ---

group_means <- soil_par_clean %>%
  group_by(Group) %>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))

print(group_means)

# --- 3. Run ANOVA for each numeric variable ---

anova_results <- soil_par_clean %>%
  select(Group, where(is.numeric)) %>%
  pivot_longer(-Group, names_to = "Variable", values_to = "Value") %>%
  group_by(Variable) %>%
  summarise(
    p_value = tryCatch({
      summary(aov(Value ~ Group))[[1]][["Pr(>F)"]][1]
    }, error = function(e) NA)
  )

print(anova_results)

# --- 4. Combine means + ANOVA p-values into one summary table ---

summary_table <- group_means %>%
  pivot_longer(-Group, names_to = "Variable", values_to = "Mean") %>%
  pivot_wider(names_from = Group, values_from = Mean) %>%
  left_join(anova_results, by = "Variable") %>%
  mutate(Significance = case_when(
    p_value < 0.001 ~ "***",
    p_value < 0.01  ~ "**",
    p_value < 0.05  ~ "*",
    TRUE ~ ""
  )) %>%
  arrange(p_value)

# View
print(summary_table)



# --- 5. Tukey HSD post-hoc tests (for significant ANOVAs) ---
tukey_results <- list()
sig_vars <- anova_results %>% filter(p_value < 0.05) %>% pull(Variable)

for (var in sig_vars) {
  model <- aov(reformulate("Group", var), data = soil_par_clean)
  tukey_results[[var]] <- TukeyHSD(model)
}

# Convert list of TukeyHSD outputs into a single data frame
tukey_table <- map_dfr(names(tukey_results), function(var) {
  # Extract the subtable (often named "Group")
  tk <- tukey_results[[var]][[1]]  # take first element (the actual comparison table)
  
  # Convert to tibble
  tk_df <- as.data.frame(tk) %>%
    rownames_to_column("contrast") %>%
    mutate(Variable = var) %>%
    rename(
      diff = diff,
      lwr = lwr,
      upr = upr,
      p_adj = `p adj`
    ) %>%
    select(Variable, contrast, diff, lwr, upr, p_adj)
  
  return(tk_df)
})

#Or combine both into one workbook with two sheets:
wb <- createWorkbook()
addWorksheet(wb, "Group Means & ANOVA")
addWorksheet(wb, "Tukey Posthoc")
writeData(wb, "Group Means & ANOVA", summary_table)
writeData(wb, "Tukey Posthoc", tukey_table)
saveWorkbook(wb, "G:/My Drive/CityDeal/Eurofins_results/Soil_full_statistical_summary.xlsx", overwrite = TRUE)


# write.csv(summary_table, "G:/My Drive/CityDeal/Eurofins_results/SUMMARY_STATS_SOIL")
```

```{r}
tukey_results <- map_dfr(
  anova_results %>% filter(p_value < 0.05) %>% pull(Variable),
  function(var) {
    model <- aov(reformulate("Group", var), data = data_clean)
    tk <- TukeyHSD(model)
    tidy(tk) %>%
      mutate(Variable = var) %>%
      select(Variable, contrast = term, group_comparison = contrast, 
             diff, lwr, upr, adj.p.value = adj.p.value)
  },
  .id = "id"
)
```

# Inclusion of soil parameters

Input files:

-   ASV table(s)

-   Metadata (soil parameters)

-   Alpha diversity data (in metadata file)

Needed packages:

```{r}
library(corrplot)
library(dplyr)
library(olsrr)
library(tidyverse)
library(sjPlot)
library(lme4)
library(car)
library(vegan)
library(adespatial)
library(ggplot2)
library(gtools)
library (ggrepel)
```

## Multi-colinearity detection

First we're going to check for (multi)-colinearity of the soil factores

### Build GLM Correlation Model

```{r}

#https://www.youtube.com/watch?v=1K3SKsEj9eI (Basic statistics)
#https://www.youtube.com/watch?v=WnmwuD8OwMw (GLM in R)
#https://www.youtube.com/watch?v=DDP62EUMRFs (GLM family) 
## Necessary libraries

```

First we need to build the correlation model. The file beneath is a file of each soil sample with for each parameter the z-score. The z-score was calculated previously in Excel with the following formula: (Average_values_parameter - Sample_value)/Standard_deviation_parameter.

rownames(my_data_corr) \<- make.unique(as.character(my_data_corr[, 1])) \# Ensure unique row names

```{r}
my_data_corr=read.csv("G:/My Drive/CityDeal/Eurofins_results/Eurofins_soil_analysis_z-score.cs", header = T, row.names = 1)
#rownames(my_data_corr) <- make.unique(as.character(my_data_corr[, 1]))  # Ensure unique row names
#my_data_corr <- my_data_corr[, -1]  # Remove the first column if it's no longer needed as data
```

### 1. CORRELATION

Correlations close to-1 or +1 might indicate the existence of multicollinearity.

As a rule of thumb, one might suspect multicollinearity when the correlation between two (predictor) variables is below -0.9 or above +0.9, though in FIMCB <!--# Ask Panji for a reference to this paper  --> paper, the cut off value is set to -0.7 or +0.7

```{r}
#Subsetting my data, to make only use the important parameters
my_data_corr_subset <- my_data_corr[, 8:ncol(my_data_corr)]

# Ensure all selected columns are numeric (optional, but good practice)
my_data_corr_subset <- my_data_corr_subset[sapply(my_data_corr_subset, is.numeric)]

cor_df <- as.data.frame(cor(my_data_corr_subset, method = 'pearson'))

corrplot(cor(my_data_corr_subset), method = "number")
```

#### Check Collinearity ----

All environmental variables were first tested for possible collinearity based on Pearson correlation values. When the correlation between two variables was higher than 0.7 (or lower than -0.7), only one of the variables was included in the downstream analyses.

```{r}
# select metadata column, containing numbers only

cor_result <- cor(as.matrix(my_data_corr_subset), method = 'pearson')
xy <- t(combn(colnames(cor_result), 2))
cor_paired_df <- data.frame(xy, dist=cor_result[xy])
# Get pairs with a cor value > 0.7
pair_to_revise <- cor_paired_df %>% filter(abs(dist) > 0.7)

pair_to_revise

## I saved the table to export from it
#write.csv(pair_to_revise, "~/Citydeal/pairs_to_revise_soildata.csv")

## Now remove data that is colinear iteratively
my_data_corr_subset$CEC = NULL

cor_result <- cor(as.matrix(my_data_corr_subset), method = 'pearson')
xy <- t(combn(colnames(cor_result), 2))
cor_paired_df <- data.frame(xy, dist=cor_result[xy])

```

After above test I manually removed certain parameters that correlate strongly with others. I this case I chose to remove phCaCl2, Dichtheid, SLV, C/S, Cavrd, Cavoorr, CECbez, C/OS, stuif, zand. I did this manually in Excel, because I wanted to check the data anyways

### 2. TOLERANCE & VARIANCE INFLATION FACTOR (VIF)

As a rule of thumb, a VIF exceeding 5 requires further investigation, whereas VIFs above 10 indicate multicollinearity. Ideally, the Variance Inflation Factors are below 3.

Preperation of data. I am planning to use the R-package car for this analysis. I subselected data from the earlier exercise.

```{r}
# This can be skipped, based on if there was colinearity in the previous step and how you solved it
my_data_after_revision <- read.csv2("~/Citydeal/Metadata/Soil_parameters_after_filter_and_alphadiv.csv", header = T, sep=",")
rownames(my_data_after_revision) <- make.unique(as.character(my_data_after_revision[, 1]))
my_data_after_revision <- my_data_after_revision[, -1]  # Remove the first column 
# Ensure all selected columns are numeric (optional, but good practice)
my_data_after_revision <- as.data.frame(lapply(my_data_after_revision, type.convert, as.is = TRUE))
my_data_after_revision[,"type"] <- as.factor(my_data_after_revision[,"type"])


```

We will test if the values are multi-colinear with different alpha diversity indexes (richness, shannon, chao), based on linear models. For this I made a document in Excel with all the Z-scores. Alpha diversity measures where found in the .rmd file for Alpha diversity.

```{r}
library(webshot)

#First we need a (generalized) linear model
my_data_after_revision <- read.csv("G:/My Drive/CityDeal/Metadata/Soil_metadata_easy_zscore.csv", row.names = 1, sep=",")

## For ITS
#shannon
Model_shannonITS<- glm(data_shannon.ITS ~ data_shannon.16S + CN + N.Levering + P.totaal + Saanv + pH+ CEC + verkr.bh + verslemp + Ca.bez + Silt, data = my_data_after_revision)
#Richness
Model_RichnessITS<- glm(RichnessITS ~ Richness.16S + CN + N.Levering + P.totaal + Saanv + pH+ CEC + verkr.bh + verslemp + Ca.bez + Silt, data = my_data_after_revision)
#Chao, because why not?
Model_chaoITS<- glm(S.chao1.ITS ~ S.chao1.16S + CN + N.Levering + P.totaal + Saanv + pH+ CEC + verkr.bh + verslemp + Ca.bez + Silt, data = my_data_after_revision)


##For 16s
#shannon
Model_shannon16s<- glm(Shannon16S ~ ShannonITS + CN + N.Levering + P.totaal + Saanv + pH+ CEC + verkr.bh + verslemp + Ca.bez + Silt, data = my_data_after_revision)
#Richness
Model_Richness16s<- glm(Richness16S ~ RichnessITS + CN + N.Levering + P.totaal + Saanv + pH+ CEC + verkr.bh + verslemp + Ca.bez + Silt, data = my_data_after_revision)
#Chao, because why not?
Model_chao16s<- glm(Chao1.16s ~ Chao1ITS + CN + N.Levering + P.totaal + Saanv + pH+ CEC + verkr.bh + verslemp + Ca.bez + Silt, data = my_data_after_revision)

## A nice table visualisation (sjPlot package)
tab_model(Model_RichnessITS)#, file ="G:/My Drive/CityDeal/ITS-Sjoerd/eurofins_soil/tab_model.html")
tab_model(Model_Richness16s, file= "G:/My Drive/CityDeal/16S-Sjoerd/eurofins_soil/tab_model.html")

webshot("G:/My Drive/CityDeal/ITS-Sjoerd/eurofins_soil/tab_model.html", "G:/My Drive/CityDeal/ITS-Sjoerd/eurofins_soil/tab_model.png")
webshot("G:/My Drive/CityDeal/16S-Sjoerd/eurofins_soil/tab_model.html", "G:/My Drive/CityDeal/16S-Sjoerd/eurofins_soil/tab_model.png")
## Non have a significant effect on Richnes (!!!!!)
```

TREES ONLY

```{r}
library(webshot)

#First we need a (generalized) linear model
my_data_after_revision <- read.csv("G:/My Drive/CityDeal/Metadata/Soil_metadata_easy_zscore_TREES.csv", row.names = 1, sep="\t")

## For ITS
## For ITS
#shannon
Model_shannonITS<- glm(data_shannon.ITS ~ data_shannon.16S + CN + N.Levering + P.totaal + Saanv + pH+ CEC + verkr.bh + verslemp + Ca.bez + Silt, data = my_data_after_revision)
#Richness
Model_RichnessITS<- glm(RichnessITS ~ Richness.16S + CN + N.Levering + P.totaal + Saanv + pH+ CEC + verkr.bh + verslemp + Ca.bez + Silt, data = my_data_after_revision)
#Chao, because why not?
Model_chaoITS<- glm(S.chao1.ITS ~ S.chao1.16S + CN + N.Levering + P.totaal + Saanv + pH+ CEC + verkr.bh + verslemp + Ca.bez + Silt, data = my_data_after_revision)


##For 16s
#shannon
Model_shannon16s<- glm(data_shannon.16S ~ data_shannon.ITS + CN + N.Levering + P.totaal + Saanv + pH+ CEC + verkr.bh + verslemp + Ca.bez + Silt, data = my_data_after_revision)
#Richness
Model_Richness16s<- glm(Richness.16S ~ Richness.ITS + CN + N.Levering + P.totaal + Saanv + pH+ CEC + verkr.bh + verslemp + Ca.bez + Silt, data = my_data_after_revision)
#Chao, because why not?
Model_chao16s<- glm(Chao1.16s ~ Chao1.ITS + CN + N.Levering + P.totaal + Saanv + pH+ CEC + verkr.bh + verslemp + Ca.bez + Silt, data = my_data_after_revision)

## A nice table visualisation (sjPlot package)
tab_model(Model_RichnessITS)#, file ="G:/My Drive/CityDeal/ITS-Sjoerd/ITS-11912/eurofins_soil/tab_model_TREES.html")

tab_model(Model_Richness16s, file= "G:/My Drive/CityDeal/16S-Sjoerd/eurofins_soil/tab_model_TREES.html")

webshot("G:/My Drive/CityDeal/ITS-Sjoerd/eurofins_soil/tab_model.html", "G:/My Drive/CityDeal/ITS-Sjoerd/eurofins_soil/tab_model_TREES.png")
webshot("G:/My Drive/CityDeal/16S-Sjoerd/eurofins_soil/tab_model.html", "G:/My Drive/CityDeal/16S-Sjoerd/eurofins_soil/tab_model_TREES.png")
```

Still i am going to test for multi-colinearity, although I don't think it matters much now. As a rule of thumb, a VIF exceeding 5 requires further investigation, whereas VIFs above 10 indicate multicollinearity. Ideally, the Variance Inflation Factors are below 3.

```{r}

## Calculate VIF score 
vif_resultsITS <- ols_vif_tol(Model_shannonITS)
vif_results16s <- ols_vif_tol(Model_shannon16s)

vif_resultsITS
vif_results16s
```

## Hellinger transformation

We need to transform the ASV table(s) to normalize them or something (check). Here we run into a little problem that is caused by the pooling of three samples for each dataset. To solve this we can do several things.

1.  We need our ASV tables

    ```{r}
    #For ITS:
    ITS_ASV_table <- read.csv("G:/My Drive/CityDeal/ITS-Sjoerd/ITS-11912/ASV_ITS_rarefied11912.csv", header = T, row.names = 1)

    #For 16s:
    ASV_table_16s <- read.csv("G:/My Drive/CityDeal/16S - Panji/Sjoerd R - 40,000 reads/phyraresuball_asv_samplenames.csv", header = T, row.names = 1)

    ```

2.  Give every sample present in the ASV table the same as the once it is pooled with.

    ```{r}
    # This file does that:
    metadata_filt <- read.csv("G:/My Drive/CityDeal/Metadata/soil_metadata_zscore_after_filtering_unpooled.csv", header = T, row.names = 1)
    ```

```{r}
# Filter out all samples that are not in metadata
filtered_ASV_16s <- t(ASV_table_16s)
filtered_ASV_16s <- filtered_ASV_16s[rownames(filtered_ASV_16s) %in% rownames(metadata_filt), ]
# See if all samples that are in the metadata are present
missing_rows_16s <- rownames(metadata_filt)[!rownames(metadata_filt) %in% rownames(filtered_ASV_16s)]
missing_rows_16s
#Now remove rows that are not in the ASV-table but are in the metadata (lost during rarefaction)
metadata_filt_16s <- metadata_filt[rownames(metadata_filt) %in% rownames(filtered_ASV_16s),]


## And for ITS
filtered_ASV_ITS <- t(ITS_ASV_table)
filtered_ASV_ITS <- filtered_ASV_ITS[rownames(filtered_ASV_ITS) %in% rownames(metadata_filt), ]
# See if all samples that are in the metadata are present
missing_rows_ITS <- rownames(metadata_filt)[!rownames(metadata_filt) %in% rownames(filtered_ASV_ITS)]
missing_rows_ITS
#Now remove rows that are not in the ASV-table but are in the metadata (lost during rarefaction)
metadata_filt_ITS <- metadata_filt[rownames(metadata_filt) %in% rownames(filtered_ASV_ITS),]





asv.16s.hell <- decostand(filtered_ASV_16s, 'hell')
asv.ITS.hell <- decostand(filtered_ASV_ITS, 'hell')


#rm(ASV_table_ITS)
#rm(filtered_ASV_ITS)
```

### Extra: tree or no

```{r}
my_data_tree <- subset(my_data_corr,subset = my_data_corr$X.1 == "with trees")
my_data_without <- subset (my_data_corr, subset = my_data_corr$X.1 == "without trees")

# Filter out all samples that are not in metadata
filtered_ASV_16s <- t(ASV_table_16s)
filtered_ASV_16s <- filtered_ASV_16s[rownames(filtered_ASV_16s) %in% rownames(metadata_filt), ]
filtered_ASV_16s_with_trees <- filtered_ASV_16s[rownames(filtered_ASV_16s) %in% rownames(my_data_tree)]
# See if all samples that are in the metadata are present
missing_rows_16s <- rownames(metadata_filt)[!rownames(metadata_filt) %in% rownames(filtered_ASV_16s)]
missing_rows_16s
#Now remove rows that are not in the ASV-table but are in the metadata (lost during rarefaction)
metadata_filt_16s <- metadata_filt[rownames(metadata_filt) %in% rownames(filtered_ASV_16s),]


## And for ITS
filtered_ASV_ITS <- t(ITS_ASV_table)
filtered_ASV_ITS <- filtered_ASV_ITS[rownames(filtered_ASV_ITS) %in% rownames(metadata_filt), ]
# See if all samples that are in the metadata are present
missing_rows_ITS <- rownames(metadata_filt)[!rownames(metadata_filt) %in% rownames(filtered_ASV_ITS)]
missing_rows_ITS
#Now remove rows that are not in the ASV-table but are in the metadata (lost during rarefaction)
metadata_filt_ITS <- metadata_filt[rownames(metadata_filt) %in% rownames(filtered_ASV_ITS),]





asv.16s.hell <- decostand(filtered_ASV_16s, 'hell')
asv.ITS.hell <- decostand(filtered_ASV_ITS, 'hell')
```

## Forward selection

I ran into the problem that my variables do not all have a significant effect on the community structure. So before running the forward selection, I am going to try to evaluate each specific variable.

```{r}
CN_mantel.ITS <- mantel(vegdist(asv.ITS.hell), dist(metadata_filt_ITS$CN))
N.lev.mantel.ITS <- mantel(vegdist(asv.ITS.hell), dist(metadata_filt_ITS$N.Levering))
P.totaal.mantel.ITS <- mantel(vegdist(asv.ITS.hell), dist(metadata_filt_ITS$P.totaal))
Saanv.mantel.ITS <- mantel(vegdist(asv.ITS.hell), dist(metadata_filt_ITS$Saanv))
ph.mantel.ITS <- mantel(vegdist(asv.ITS.hell), dist(metadata_filt_ITS$pH))
CEC.mantel.ITS <- mantel(vegdist(asv.ITS.hell), dist(metadata_filt_ITS$CEC))
verkr.bh.mantel.ITS <- mantel(vegdist(asv.ITS.hell), dist(metadata_filt_ITS$verkr.bh))
verslemp.mantel.ITS <- mantel(vegdist(asv.ITS.hell), dist(metadata_filt_ITS$verslemp))
Ca.bez.mantel.ITS <- mantel(vegdist(asv.ITS.hell), dist(metadata_filt_ITS$Ca.bez))
silt.mantel.ITS <- mantel(vegdist(asv.ITS.hell), dist(metadata_filt_ITS$Silt))
```

```{r}
CN_mantel.16s <- mantel(vegdist(asv.16s.hell), dist(metadata_filt_16s$CN))
N.lev.mantel.16s <- mantel(vegdist(asv.16s.hell), dist(metadata_filt_16s$N.Levering))
P.totaal.mantel.16s <- mantel(vegdist(asv.16s.hell), dist(metadata_filt_16s$P.totaal))
Saanv.mantel.16s <- mantel(vegdist(asv.16s.hell), dist(metadata_filt_16s$Saanv))
ph.mantel.16s <- mantel(vegdist(asv.16s.hell), dist(metadata_filt_16s$pH))
CEC.mantel.16s <- mantel(vegdist(asv.16s.hell), dist(metadata_filt_16s$CEC))
verkr.bh.mantel.16s <- mantel(vegdist(asv.16s.hell), dist(metadata_filt_16s$verkr.bh))
verslemp.mantel.16s <- mantel(vegdist(asv.16s.hell), dist(metadata_filt_16s$verslemp))
Ca.bez.mantel.15s <- mantel(vegdist(asv.16s.hell), dist(metadata_filt_16s$Ca.bez))
silt.mantel.16s <- mantel(vegdist(asv.16s.hell), dist(metadata_filt_16s$Silt))
```

A check for ITS

```{r}
CN_mantel.ITS
N.lev.mantel.ITS
P.totaal.mantel.ITS
Saanv.mantel.ITS
ph.mantel.ITS
CEC.mantel.ITS
verkr.bh.mantel.ITS
verslemp.mantel.ITS
Ca.bez.mantel.ITS
silt.mantel.ITS
```

And a check for 16s

```{r}
CN_mantel.16s
N.lev.mantel.16s
P.totaal.mantel.16s
Saanv.mantel.16s
ph.mantel.16s
CEC.mantel.16s
verkr.bh.mantel.16s
verslemp.mantel.16s
Ca.bez.mantel.16s
silt.mantel.16s
```

Now it's time for forward selection. We do this to know which variables we should keep for the RDA analysis.

```{r}
sel.fs <- forward.sel(asv.ITS.hell, metadata_filt_ITS, nperm = 999)
pval.adj <- p.adjust(sel.fs$pval, method = 'holm', n = ncol(metadata_filt_ITS))
sel.fs$pval.adj <- pval.adj
sel.fs$pval.adj.stars <- gtools::stars.pval (pval.adj)
sel.fs
sel.fs.ITS <- sel.fs

```

And for 16S

```{r}
sel.fs <- forward.sel(asv.16s.hell, metadata_filt_16s, nperm = 999)
pval.adj <- p.adjust(sel.fs$pval, method = 'holm', n = ncol(metadata_filt_16s))
sel.fs$pval.adj <- pval.adj
sel.fs$pval.adj.stars <- gtools::stars.pval (pval.adj)
sel.fs
sel.fs.16s <- sel.fs
sel.fs.16s
```

### The RDA analysis for 16S.

```{r}
# RDA analysis ------------------------------------------------------------

#If I want to make a subselection
#metadata_select_16s <- metadata_filt_16s %>% select(Silt)

tb_rda.all <- rda(formula = asv.16s.hell ~ CN + N.Levering + P.totaal + pH + CEC + verslemp + Ca.bez + Silt, data = metadata_filt_16s)

anova(tb_rda.all)  # 0.001 *** - it is significant


```

```         
Permutation test for rda under reduced model Permutation: free Number of permutations: 999  Model: rda(formula = asv.16s.hell ~ CN + N.Levering + P.totaal + pH + CEC + verslemp + Ca.bez + Silt, data = metadata_filt_16s)           Df Variance      F Pr(>F)     Model      8  0.04820 2.5429  0.001 *** Residual 288  0.68235                   --- Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
```

##Permutation test for rda under reduced model Permutation: free Number of permutations: 999

Model: rda(formula = asv.16s.hell \~ CN + N.Levering + P.totaal + Saanv + pH + CEC + verkr.bh + verslemp + Ca.bez + Silt, data = metadata_filt_16s) Df Variance F Pr(\>F)\
Model 10 0.05347 2.2584 0.001 ***Residual 286 0.67708\
--- Signif. codes: 0 ‘***’ 0.001 ‘\*\*’ 0.01 ‘\*’ 0.05 ‘.’ 0.1 ‘ ’ 1

```{r}
# tb_rda.all.ITS <- rda(formula = asv.ITS.hell ~ Silt + Saanv, data = metadata_filt_ITS)
# 
# anova(tb_rda.all.ITS)  # 0.001 *** - it is significant

tb_rda.all.ITS <- rda(formula = asv.ITS.hell ~ Silt, data = metadata_filt_ITS)
anova(tb_rda.all.ITS)  # 0.001 *** - it is significant
```

```         
Permutation test for rda under reduced model Permutation: free Number of permutations: 999  Model: rda(formula = asv.ITS.hell ~ Silt + Saanv, data = metadata_filt_ITS)           Df Variance      F Pr(>F)     Model      2  0.01083 1.6206  0.001 *** Residual 274  0.91544                   --- Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
```

```{r}
adjR2.tbrda.16s <- RsquareAdj(tb_rda.all)$adj.r.squared
adjR2.tbrda.16s
```

```{r}
adjR2.tbrda.ITS <- RsquareAdj(tb_rda.all.ITS)$adj.r.squared
adjR2.tbrda.ITS
```

```{r}
rda_plot <- plot(tb_rda.all.ITS, scaling=1)
rda_plot
```

## Plot RDA with ggplot2

### Normal RDA plot

Let's try something. This is fun, but I should add color based on typology

### RDA - biplot16S

```{r}
R.sum <- summary(tb_rda.all)
tb_rda.all$CA$eig
R.sum   # Prints the "Importance of components" table
R.sum$cont$importance[2, "RDA1"]
# 0.74785
R.sum$cont$importance[2, "RDA2"]
# [1] 0.19804
```

with help from chatGPT

```{r}
library(ggplot2)
library(ggrepel)


# Extract eigenvalues and percentage variance explained
eig_vals <- tb_rda.all$CCA$eig  
axis_perc <- eig_vals / sum(eig_vals) * 100

# Labels for axes
xlab_text <- paste0("RDA1 (", round(axis_perc[1], 1), "%)")
ylab_text <- paste0("RDA2 (", round(axis_perc[2], 1), "%)")

# Base RDA plot with metadata aesthetics
rda_plot <- ggplot() + 
  geom_vline(xintercept = 0, color = "black", linetype = 2) +
  geom_hline(yintercept = 0, color = "black", linetype = 2) +  
  xlab(xlab_text) +
  ylab(ylab_text) +
  scale_x_continuous(sec.axis = dup_axis(labels = NULL, name = NULL)) +
  scale_y_continuous(sec.axis = dup_axis(labels = NULL, name = NULL)) +

  # 🎨 Sample points
  geom_point(data = sites.annotated, 
             aes(x = axis1, y = axis2, fill = topology, shape = city), 
             size = 5) +

  # # 🔴 Species scores (smaller red dots)
  # geom_text(data = species.long2, 
  #           aes(x = axis1, y = axis2, label = labels), 
  #           color = "red", size = 3) +

  # 🌿 Environmental arrows from envfit
  geom_segment(data = arrow_df, 
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2), 
               arrow = arrow(length = unit(0.2, "cm")), 
               color = "blue", size = 0.7) +

  # 🏷 Environmental labels
  geom_text_repel(data = arrow_df, 
                  aes(x = RDA1, y = RDA2, label = labels), 
                  color = "blue", size = 4) +

  # Optional color palette for typology
  scale_fill_manual(values = c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECC8')) +
   scale_shape_manual(values=c(21,22,23,24,25)) +

  # Style
  mytheme
rda_plot

ggsave("G:/My Drive/CityDeal/16S-Sjoerd/eurofins_soil/RDA_environement_16S.png",width = 15, height = 10, units = "in", dpi = 600)


```

And for ITS

```{r}
R.sum <- summary(tb_rda.all.ITS)
tb_rda.all$CA$eig
R.sum   # Prints the "Importance of components" table
R.sum$cont$importance[2, "RDA1"]
# 0.74785
R.sum$cont$importance[2, "PC1"]
# [1] 0.19804
```

```{r}
plot2 <- ordiplot(tb_rda.all.ITS, choices=c(1,2))
sites.long2 <- BiodiversityR::sites.long(plot2, env.data=asv.ITS.hell) #to map the position of each sample on RDA plot
species.long2 <- BiodiversityR::species.long(tb_rda.all.ITS) #to map the position of each ASV on RDA plot
axis.long2 <- BiodiversityR::axis.long(plot2, choices=c(1, 2)) #to quantify axis on RDA plot

sites.long2$axis1 = as.numeric(sites.long2$axis1) #change the parameter to numeric
sites.long2$axis2 = as.numeric(sites.long2$axis2) #change the parameter to numeric
metadata_per_sample$sample <- rownames(metadata_per_sample)
sites.long2$sample<-rownames(sites.long2)
sites_annotated <- merge(sites.long2, metadata_per_sample, by = "sample")

sites.long2 <- sites.annotated

# Run envfit on your RDA object and metadata
spec.envfit <- envfit(tb_rda.all.ITS, env = metadata_filt_ITS)

# Convert the arrow output to a data frame for ggplot
selected_vars <- rownames(tb_rda.all.ITS$CCA$biplot)
arrow_df <- as.data.frame(spec.envfit$vectors$arrows) %>%
  tibble::rownames_to_column("labels") %>%
  mutate(
    r = spec.envfit$vectors$r,
    p = spec.envfit$vectors$pvals
  )
arrow_df <- arrow_df %>%
  filter(labels %in% selected_vars)

# Extract eigenvalues and percentage variance explained
eig_vals <- tb_rda.all.ITS$CCA$eig  
axis_perc <- eig_vals / sum(eig_vals) * 100

# Labels for axes
xlab_text <- paste0("RDA1 (", round(axis_perc[1], 1), "%)")
ylab_text <- paste0("PC1") #(", round(axis_perc[2], 1), "%)")

# Base RDA plot with metadata aesthetics
rda_plot <- ggplot() + 
  geom_vline(xintercept = 0, color = "black", linetype = 2) +
  geom_hline(yintercept = 0, color = "black", linetype = 2) +  
  xlab(xlab_text) +
  ylab(ylab_text) +
  scale_x_continuous(sec.axis = dup_axis(labels = NULL, name = NULL)) +
  scale_y_continuous(sec.axis = dup_axis(labels = NULL, name = NULL)) +

  # 🎨 Sample points
  geom_point(data = sites_annotated, 
             aes(x = axis1, y = axis2, fill = topology, shape = city), 
             size = 5) +

  # # 🔴 Species scores (smaller red dots)
  # geom_text(data = species.long2, 
  #           aes(x = axis1, y = axis2, label = labels), 
  #           color = "red", size = 3) +

  # 🌿 Environmental arrows from envfit
  geom_segment(data = arrow_df, 
               aes(x = 0, y = 0, xend = RDA1, yend = PC1), 
               arrow = arrow(length = unit(0.2, "cm")), 
               color = "blue", size = 0.7) +

  # 🏷 Environmental labels
  geom_text_repel(data = arrow_df, 
                  aes(x = RDA1, y = PC1, label = labels), 
                  color = "blue", size = 4) +

  # Optional color palette for typology
  scale_fill_manual(values = c('#FFABAB', '#ACE7FF', '#DBFFD6', '#ECD4FF', '#FEECC8')) +
   scale_shape_manual(values=c(21,22,23,24,25)) +

  # Style
  mytheme

rda_plot

ggsave("G:/My Drive/CityDeal/ITS-Sjoerd/ITS-11912/eurofins_soil/RDA_environement_ITS_Siltonly.png",width = 15, height = 10, units = "in", dpi = 600)

```

other plots

```{r}
# Base plot for R² and Adjusted R² over steps
ggplot(sel.fs, aes(x = factor(order), group = 1)) +
  geom_line(aes(y = R2Cum, color = "R² Cumulative"), size = 1.2) +
  geom_line(aes(y = AdjR2Cum, color = "Adjusted R² Cumulative"), size = 1.2, linetype = "dashed") +
  geom_point(aes(y = R2Cum), size = 3, shape = 21, fill = "blue") +
  geom_point(aes(y = AdjR2Cum), size = 3, shape = 21, fill = "red") +
  scale_color_manual(values = c("R² Cumulative" = "blue", "Adjusted R² Cumulative" = "red")) +
  labs(x = "Variable Addition Order",
       y = "Cumulative R² / Adjusted R²",
       color = "Legend",
       title = "Model Improvement Over Forward Selection Steps")
  
```

```{r}
# Bar plot of R² contributions
ggplot(sel.fs, aes(x = reorder(variables, -R2), y = R2)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "Variable", y = "Incremental R²",
       title = "Incremental Contribution of Each Variable (ΔR²)") +
  theme_minimal() +
  coord_flip()
```
